RAG suppresses group 2 innate lymphoid cells

Version 1

Abstract

Abstract

Antigen specificity is the central trait distinguishing adaptive from innate immune function. Assembly of antigen-specific T cell and B cell receptors occurs through V(D)J recombination mediated by the Recombinase Activating Gene endonucleases RAG1 and RAG2 (collectively called RAG). In the absence of RAG, mature T and B cells do not develop and thus RAG is critically associated with adaptive immune function. In addition to adaptive T helper 2 (Th2) cells, group 2 innate lymphoid cells (ILC2s) contribute to type 2 immune responses by producing cytokines like Interleukin-5 (IL-5) and IL-13. Although it has been reported that RAG expression modulates the function of innate natural killer (NK) cells, whether other innate immune cells such as ILC2s are affected by RAG remains unclear. We find that in RAG-deficient mice, ILC2 populations expand and produce increased IL-5 and IL-13 at steady state and contribute to increased inflammation in atopic dermatitis (AD)-like disease. Further, we show that RAG modulates ILC2 function in a cell-intrinsic manner independent of the absence or presence of adaptive T and B lymphocytes. Lastly, employing multiomic single cell analyses of RAG1 lineage-traced cells, we identify key transcriptional and epigenomic ILC2 functional programs that are suppressed by a history of RAG expression. Collectively, our data reveal a novel role for RAG in modulating innate type 2 immunity through suppression of ILC2s.

Introduction

Atopic disorders such as atopic dermatitis (AD), asthma, and food allergy are associated with T helper type 2 (Th2) cell responses, elevated production of the type 2 cytokines interleukin(IL)-4, IL-5, and IL-13, and induction of immunoglobulin(Ig)E1–4. Classically, this allergic inflammatory cascade is believed to originate with antigenic stimulation of T cell receptors on adaptive T cells, which in turn results in the production of IgE from B and plasma cells capable of binding the same antigen. Indeed, presence of antigen-specific IgE reactivity is a hallmark across atopic disorders.5 Thus, for decades, antigen-specific adaptive Th2 cell responses have been the primary focus of investigation in the pathogenesis of atopic diseases. However, recent studies indicate that innate immune cells are sufficient to not only drive allergic pathology, but also amplify adaptive Th2 cell responses6–9. These studies suggest that innate immune mechanisms may play a larger role in driving atopic inflammation than previously recognized.

Innate lymphoid cells (ILCs), while lacking antigen receptors generated by recombination activating gene proteins RAG1 and RAG2 (collectively called RAG), are the innate counterparts of T cells. For example, ILC2s mirror adaptive Th2 cells in their developmental requirements, cytokine profiles, and effector functions10. Unlike classical T cells, ILC2s are concentrated at barrier surfaces to rapidly respond to microbial and environmental stimuli. ILC2s are key mediators of inflammatory skin conditions like AD11–13. Indeed, in murine models of AD-like disease, type 2 skin inflammation can still occur despite the absence of adaptive T cells, but is further reduced after depletion of ILC2s12,13. Furthermore, recent studies have shown that ILC2s harbor non-redundant functions in the presence of the adaptive immune system in the setting of anti-helminth immunity14,15. These findings suggest that ILC2 dysfunction may also uniquely contribute to the pathogenesis of atopic diseases, independent of adaptive immunity. However, the cell-intrinsic mechanisms that drive ILC2 dysregulation remain poorly understood.

ILC2s were originally discovered due to their capacity to orchestrate multiple allergic pathologies in immunocompromised mice, most notably in RAG-deficient mice that lack T and B cells16–22. These discoveries fundamentally redefined our understanding of allergic diseases and placed a major focus on ILC2s as potential drivers of human allergic disease. However, despite ILC2s not requiring RAG expression for their development, fate mapping studies in mice have demonstrated that up to 60% of ILC2s have historically expressed RAG1 during development23,24. Although previous work has described roles of RAG beyond antigen receptor recombination in developing T and B cells25–27 and NK cells24, how this developmental expression of RAG impacts ILC2s remains unclear.

By directly comparing RAG-deficient and RAG-sufficient mice, we unexpectedly found enhanced AD-like disease in RAG-deficient mice, despite the lack of adaptive lymphocytes to contribute to AD-like inflammation. Using splenocyte replenishment and bone marrow chimeras, we show that RAG suppresses ILC2 activation and expansion in a cell-intrinsic manner. Employing a RAG1-lineage reporter mouse line, we performed simultaneous single-cell multiomic RNA and ATAC sequencing to show that RAG fate-mapped ILC2s display unique transcriptional and epigenomic alterations consistent with the suppression of effector cytokine production. Collectively, our studies reveal evolutionarily conserved regulatory functions of RAG within innate lymphocytes, extending beyond the generation of antigen receptors in adaptive lymphocytes.

Results

RAG deficiency leads to expansion and activation of ILC2s

AD-like disease can be elicited in the skin of mice with repeated application of the topical vitamin D analog calcipotriol (MC903)28. Although it has been previously demonstrated that MC903 can induce AD-like disease in RAG-deficient mice that lack T and B cells, in part via ILC2 activation12,13, the relative contributions of ILC2s and the adaptive lymphocyte compartment have not been rigorously evaluated. We hypothesized that the presence of Th2 cells, in addition to ILC2s, would lead to enhanced AD-like disease in an additive fashion. In testing this, we evaluated both RAG1-sufficient wild-type (WT) mice and RAG1-deficient Rag1−/− mice in the setting of AD-like disease (Fig. 1A). Unexpectedly, we observed that Rag1−/− mice developed increased ear skin thickness (Fig. 1B) and higher frequencies of ILC2s in the skin-draining lymph nodes (sdLNs) compared to control WT mice (Fig. 1C, S1A,B). Furthermore, a larger proportion of ILC2s from Rag1−/− mice exhibited production of both IL-5 (Fig. 1D, S1C) and IL-13 (Fig. 1E, S1C). Our findings indicated that RAG1 deficiency results in paradoxically worse AD-like disease in association with enhanced ILC2 expansion and activation.

RAG deficiency leads to expansion and activation of ILC2s during inflammation and at steady state.

A) Experimental schematic of AD-like disease. WT (Control) B6 mice or Rag1−/− mice treated topically to the inner surface of each ear with 2 nmol MC903 in 10 μL ethanol vehicle daily for 7 days develop AD-like inflammation. B) Ear thickness measured daily in AD-like inflammation. Data representative of at least 2 independent experiments, 5 mice/group. ** P < 0.01 by 2-way ANOVA with Sidak’s multiple comparisons test, day 7. C) Proportion of CD90+, Lin− cells (Lin− defined as CD3−, CD5−, CD11b−, CD11c−, CD19−, NK1.1−, and FcεR1−) determined to be ILC2s (IL-33R+) in skin-draining lymph nodes (sdLN) from WT or Rag1−/− mice with AD-like ear inflammation. Percent ILC2 from sdLN in AD-like disease following PMA/iono stimulation positive for D) IL-5 or E) IL-13 staining. F) Schematic of steady state analysis of sdLN from WT (Control) or Rag1−/− mice. G) Proportion of steady state sdLN CD90+, Lin− cells determined to be ILC2s as in (C). Percent ILC2 from sdLN in steady state following PMA/iono stimulation positive for H) IL-5 or I) IL-13 staining. C-E; G-I) Data representative of at least 2 independent experiments, 4-5 mice/group * P < 0.05, ** P < 0.01, *** P < 0.001 by two-tailed Welch’s t test. All data represented as mean with standard deviation.

To determine whether this phenomenon was specific to AD-like pathological conditions, we next examined the sdLNs in Rag1−/− and lymphocyte-sufficient Rag+/- littermate control mice in the absence of disease (Fig. 1F). We found that the frequency of ILC2s was increased at steady state in Rag1−/− sdLNs (Fig. 1G) and that a higher proportion of these ILC2s produced both IL-5 (Fig. 1H) and IL-13 (Fig. 1I) compared to WT controls. The RAG recombinase requires both RAG1 and RAG2 components to successfully rearrange a functional antigen receptor in adaptive lymphocytes29. Thus, to test whether our findings are specific to RAG1, or related to function of the overall RAG complex, we similarly examined the steady-state profile of ILC2s in Rag2−/− mice (Fig. S2A). Deficiency of RAG2 led to an expansion of ILC2s in the sdLNs (Fig. S2B) and increased proportions of ILC2s expressing IL-5 (Fig. S2C) and IL-13 (Fig. S2D) similar to Rag1−/−mice. Collectively, these findings suggest that the RAG recombinase modulates ILC2 function at steady state and during type 2 inflammation. However, whether the hyperactive ILC2 phenotype is due to a cell-intrinsic process or simply due to the absence of T and B cells was unclear.

ILC2 suppression by RAG is cell intrinsic

Given the importance of the adaptive lymphocyte compartment in shaping the secondary lymphoid organ repertoire, we next wanted to examine whether the presence of adaptive lymphocytes could restore ILC2 homeostasis in RAG-deficient mice. To test this, we created splenocyte chimera mice by reconstituting both Rag1−/−and control WT mice with splenocytes containing T and B cells from WT donor mice (Fig. 2A). We first assessed the overall level of immune reconstitution in the recipient mice and found fully restored proportions of CD4+ (Fig. S3A) and CD8+ (Fig. S3B) T cells in the spleens of recipient Rag1−/− mice, although B cell numbers remained significantly lower than in WT mice (Fig. S3C). Upon induction of AD-like disease, we found that the Rag1−/−mice still exhibited increased ear skin thickness (Fig. 2B), enhanced expansion of ILC2s (Fig. 2C), and increased proportions of ILC2s expressing IL-5 (Fig. 2D) and IL-13 (Fig. 2E) in the sdLNs. Interestingly, we found significantly higher proportions of eosinophils in the spleens of Rag1−/− recipient mice (Fig. S3D), possibly reflecting the increased IL-5 production we observed from ILC2s. These findings indicate that the mere introduction of exogenous T and B cells is not sufficient to suppress ILC2 dysregulation in the setting of RAG deficiency.

Homeostatic expansion and activation of RAG-deficient ILC2s is cell intrinsic

A) Experimental schematic of AD-like disease in splenocyte chimera experiment. WT B6 or Rag1−/−mice received WT splenocytes and developed AD-like inflammation after subsequent topical treatment with 2 nmol MC903 in 10 μL ethanol vehicle to each ear daily for 10 days. B) Ear thickness measured daily in AD-like inflammation. Data representative of 2 independent experiments, 4-5 mice per group. **** P <0.0001 by 2-way ANOVA with Sidak’s multiple comparisons test, day 10. C) Proportion of CD90+, Lin− cells (Lin− defined as CD3−, CD5−, CD11b−, CD11c−, CD19−, NK1.1−, and FcεR1−) determined to be ILC2s (IL-33R+). Percent ILC2 from sdLN in splenocyte chimera mice with AD-like disease after PMA/iono stimulation positive for D) IL-5 or E) IL-13 staining. F) Schematic of bone marrow chimera experiment. Equal quantities of bone marrow cells from Rag1−/− (CD45.2, CD90.2 – orange) and WT (CD45.2, CD90.1 – blue) C57Bl/6J donor mice were used to reconstitute the immune systems of irradiated recipient WT (CD45.1 – black) C57Bl/6J mice. G) Proportion of donor (CD45.2+) ILC2 defined as in (C) in sdLN by donor source (CD90.1+ – WT, CD90.2− – Rag1−/−). Proportion of Lin− ILCs by donor source positive for H) IL-5 and I) IL-13 following PMA/iono stimulation and cytokine staining. C-E) Data representative of at least 2 independent experiments, 4-5 mice per group. ** P < 0.01, **** P < 0.0001 by two-tailed Welch’s t test. G-H) Data representative of at least 2 independent experiments, 4-5 mice per group. ** P < 0.01 by two-tailed ratio means paired t test. All data represented as mean with standard deviation.

To further test whether this phenotype is mediated by cell-intrinsic RAG expression, we next generated mixed bone marrow (BM) chimeras. We harvested BM from congenic CD90.1+ WT and CD90.2+ Rag1−/− donor mice on a CD45.2+ background and reconstituted sub-lethally irradiated CD45.1+ congenic WT recipients with a 50:50 mixture of WT:Rag1−/− BM (Fig. 2F). After confirming reconstitution of donor immune cells in the sdLN (Fig. S4A,B), we examined the frequency and activity of ILC2s in the sdLNs based on whether they originated from WT (CD90.1+) or Rag1−/−(CD90.2+) donors (Fig. S4C). Strikingly, of the total donor ILC2s, the majority were derived from Rag1−/− donors (Fig. 2G). This was not due to differences in overall donor reconstitution, since measuring all Lin− cells revealed WT donor cells outnumbered those from Rag1−/− donors (Fig. S4D). Of the total IL-5– (Fig. 2H) and IL-13-expressing (Fig. 2I) ILCs, the majority were also derived from Rag1−/−donors. Taken together, these data suggest that cell-intrinsic RAG activity in ILC2s may limit their capacity to expand and become activated.

A history of RAG expression marks a subpopulation of ILC2s in the skin draining lymph node

In contrast to resting naïve T cells, ILC2s resemble activated Th2 cells at steady state based on their transcriptomic and epigenomic profiles30,31. While both T cells and ILC2s exhibit historical RAG expression23, they do not actively express the protein in their mature state32. Taken together, these findings provoke the hypothesis that ILC2s are regulated by RAG early in development to imprint alterations that influence their activity as mature cells. To distinguish ILC2s as either having a history of RAG expression or not, we utilized a RAG lineage tracing system, whereby a Rag1Cremouse was crossed to a reporter mouse expressing tandem dimer red fluorescent protein (tdRFP) in a Cre-dependent manner from the Rosa26 locus (Fig. 3A)24,33. This system allowed us to compare RAG-experienced (RAGexp) and RAG-naïve (RAGnaïve) lymphoid cells, including ILC2s, simultaneously originating from the same immunocompetent host, thus removing confounders inherent in knockout and chimera experiments. Analysis of sdLN from the reporter mice revealed that nearly all CD4+ T cells, CD8+ T cells, and B220+ B cells expressed tdRFP (positive history of Cre expression from the Rag1 locus), consistent with the known requirement of RAG expression for their development (Fig. 3B-D,G). We also examined NK cells, since certain subsets of NK cells are known to express RAG during their development24, and we observed that 60% of NK cells were tdRFP+, similar to previous findings (Fig. 3E,G)24. In the ILC2 population of the sdLN, around 50% were tdRFP+ (Fig. 3F,G), similar to proportions of RAG fate-mapped ILC2s previously observed in the fat24 and lung23,24. These findings demonstrate that there are heterogeneous populations of ILC2s marked by differential tdRFP+ fate mapping. Importantly, this provided us with the possibility to profile these different ILC2 subsets based on Rag1Cre-activated expression of tdRFP.

A history of RAG expression marks a population of ILC2s in the sdLN.

A) Schematic of RAG fate mapping in the lymphoid cell compartment using reporter mice expressing Cre-inducible tandem dimer red fluorescent protein (tdRFP) from the Rosa26 locus crossed to mice expressing Cre recombinase from the Rag1 locus. B-F) Histograms of tdRFP signal in CD45+ sdLN cells by cell type for B) CD4+ T cells (B220−, CD3+, CD4+), C) CD8+ T cells (B220−, CD3+, CD8+), D) B cells (MHCII+, B220+), E) NK cells (B220−, CD3−, CD4−, CD8−, CD49b+, NK1.1+), F) ILC2s (B220−, CD3−, CD4−, CD8−, CD49b−, NK1.1−, CD11b−, CD11c−, SiglecF−, CD90+, KLRG1+ or ICOS+ or IL-33R+), F) quantification of tdRFP+ proportion of each cell type.

Multiomic profiling enhances the detection of rare tissue-specific ILC2s

Transient RAG expression early in lymphoid development leads to well-characterized, durable effects on B and T cell development and function mainly through successful genomic rearrangement of antigen receptors. Yet, our data indicate that RAG expression also imprints phenotypic changes on ILC2s, which can develop independently of functional antigen receptors, provoking the hypothesis that RAG expression may affect broader epigenomic and transcriptional programs. Furthermore, our data indicate that the impact of RAG on ILC2 function has implications for AD-like skin inflammation, suggesting a persistent effect that modulates phenotypes of type 2 inflammation.

To test these hypotheses, we performed combined single-nuclei RNA sequencing (snRNA-seq) and ATAC sequencing (snATAC-seq) of sdLN cells from RAG fate-mapped mice at steady state and in the setting of AD-like disease (Fig. S5A,B). Because fluorescence-activated cell sorting (FACS) can cause physical stress, cell loss, and contamination, which can introduce unwanted perturbations in target cells, instead we utilized gentle initial negative selection by magnetic activated cell sorting (MACS) to remove most B and T cells and monocytes prior to sequencing (Fig. S5A). This allowed us to enrich for innate immune cell populations prior to sequencing. Further, we used the gene encoding tdRFP as a barcode to differentiate between RAGexp (RAG fate map-positive) and RAGnaïve (RAG fate map-negative) ILC2s at the single-cell level (Fig. 5A). The multiomic data was analyzed using recently developed pipelines in Cell Ranger, Seurat34–36, and Signac37, and sequenced cells were further filtered computationally to enrich for ILCs, as in previous studies (see methods)38.

In addition to gene expression (GEX) information derived from snRNA-seq (Fig. 4A), we calculated a “gene activity” (GA) score based on chromatin accessibility at gene loci (Fig. 4B) from the corresponding snATAC-seq dataset37. Clustering the cells with each data subset alone and in combination using weighted nearest neighbor (WNN) analysis, we identified six clusters (Fig. 4C) that demonstrated consistent differences in cellular markers based on both metrics of GEX and GA (Fig. 4D,E; Table S1). Additionally, top markers for each cluster clearly differentiated each cell type (Fig. 4F). Despite successful ILC2 enrichment via MACS depletion for lineage markers and computational filtering (see methods), our data set included non-ILC2 populations determined by gene expression to be T cells, dendritic cells, B cells, and NK cells (Fig. 4A,F; Table S1), which allowed for broader multidimensional comparisons while studying this ILC2-enriched data set.

Multiomic analysis of ILC2s through single nuclei sequencing of the sdLN.

A) Schematic of the gene expression (GEX) assay derived from snRNA-seq data. B) Schematic of the gene activity (GA) assay, representing estimated transcription scores derived from snATAC-seq data using Signac. C) UMAP visualizations of independent analyses of RNA-seq and ATAC-seq data for 2034 sdLN cells after dimensional reduction and clustering combined using weighted nearest neighbor (WNN) analysis in Seurat. Cluster identities are color coded consistently throughout the following panels. Heatmaps of D) top 25 GEX marker genes and E) top 25 GA marker genes identified for each cluster. See Table S1 for full lists of genes. F) Dotplots comparing selected marker genes for each cluster between the GEX and GA assays, with emphasis on known cell type-specific markers. G) Schematic of differentially accessible (DA) chromatin assay, which finds the nearest gene to any peak calculated to be differentially open in a particular cell cluster. See Table S2 for full lists of top 25 DA cluster markers. H) Overlap of top 100 markers for the ILC2 cluster from the GEX, GA, and DA assays. See Table S3 for top 100 DA peaks and distances to nearest genes and Table S4 for full list of top 100 ILC2 markers. I) Selected genes from the ILC2 gene set for each assay individually and for overlaps.

To further complement the GEX and GA assays, we utilized another method of detecting cell-specific marker genes, whereby chromatin regions that are differentially accessible (DA), or open, in each cluster could be linked by their physical proximity to specific genes (Fig. 4G, S6; Table S2; see methods). Comparing the top 100 GEX, GA, and DA markers in the ILC2 population, we identified a multiomic ILC2 signature of 235 unique genes (Fig. 4H; Table S3,4). While there was some overlap between each respective set, the multiomic approach enabled more extensive identification of an ILC2 gene program than through either snRNA-seq or snATAC-seq alone. The analysis revealed a variety of canonical ILC2-associated genes specific to the ILC2 cluster (Fig. 4I, S6,7) including the ILC2-activating cell surface receptors Icos39,40, Il2ra41,42, Il18r138,43,44, Nmur14,15,45–47, and Il1rl1 (encoding the receptor for IL-33)11,12,48,49, as well as transcription factors such as Gata319,50–53, Bcl11b54–57, Maf19,58,59, Ets160, and Rora38,51,61–63, all previously shown to be important for ILC2 development and/or function.

Expression of some secreted proteins can be difficult to capture in droplet-based snRNA-seq experiments due to low transcript levels and relatively shallow sequencing depth. With the addition of the complementary GA and DA assays from snATAC-seq, our analysis identified Il5 (Fig. 4I, S6B), a canonical ILC2 cytokine, in the DA assay, while in the GA assay we found Bmp764,65, which has been shown to be secreted by ILC2s to influence browning of adipose tissue. Additionally, we identified the secreted chemokine Ccl1 as an ILC2 marker gene66–69, which along with its cognate receptor Ccr8 (also an ILC2 marker in our analysis)44,67,70, participates in a feed-forward circuit to drive ILC2 recruitment and expansion71. Thus, our findings demonstrate how genetic barcoding, combining transcriptomic and epigenomic analyses, and cross-validation across many published studies can yield new insights while providing internal control measures to elevate the rigor, robustness, and confidence of identifying gene signatures of rare populations such as ILC2s at the single-cell level.

ILC2s with a history of RAG expression are epigenomically suppressed

As noted above, barcoding the ILC2s afforded the opportunity to transcriptionally and epigenomically profile ILC2s under identical developmental conditions by dividing the ILC2 cluster into RAGexp and RAGnaïve populations. We hypothesized that RAGexp ILC2s would have a distinct transcriptional profile compared to ILC2s without any history of RAG expression. To test this, we calculated differentially expressed genes (DEGs) for the ILC2 cluster by Rag1 fate-map status. Genes with higher expression in RAGexp cells relative to RAGnaïve cells had positive fold change values, and vice versa, with genes relatively increased in RAGnaïve cells having negative values (Fig. S8A, Table S5). Using gene set enrichment analysis (GSEA)72,73 on the ranked list of DEGs, we found that gene sets generally representing lymphocyte activation and differentiation were suppressed in RAGexp ILC2s compared to RAGnaïve ILC2s (Fig. S8B,C, Table S6), consistent with our previous observations that ILC2s are expanded and more activated in RAG-deficient mice relative to WT mice.

We next employed newly described methodologies37,74 that quantify associations between open chromatin peaks and the expression of nearby genes to describe the functional regulomes of both RAGexp and RAGnaïve ILC2s (Fig. 5B). In this analysis, each ATAC peak can be linked to multiple genes, and each gene to multiple peaks, generating a list of “gene-to-peak links” or GPLs (see methods). For each gene, we interpreted the number of corresponding GPLs as a quantitative representation of the regulome activity for that gene. Considering RAGexp and RAGnaïve cells as two separate populations, we generated two lists of GPLs (Table S7) defining functional regulomes for each population. We focused our analysis on the functional regulomes of ILC2s by filtering the GPL lists for the 235 unique ILC2 genes identified in our multiomic gene set (Fig. 4H; Table S8), then calculated the difference in GPLs between RAGexp and RAGnaïve cells for each gene and sorted them; genes displaying greater numbers of GPLs in the RAGnaïve population are at the top, and genes with more GPLs in the RAGexp population are at the bottom (Fig. 5C; Table S9).

A history of RAG expression imprints transcriptomic and epigenomic modulation of ILC2 gene programs.

A) Schematic of transcriptional RAG fate mapping. Sequenced cells from the RAG fate map mouse (see Figure 3A) transcribe tdRFP only after Cre is expressed from the Rag1 locus. Cells were assigned as either having a history of RAG expression (RAGexp – tomato red) or not (RAGnaïve – dark gray) based on detection of tdRFP transcript in the RNA-seq data. B) Schematic of mapping gene to peak links (GPLs). The LinkPeaks function of Signac (see methods) calculates significant correlations between open chromatin at defined peaks (teal bars) and nearby gene expression. These links represent inferred epigenomic-transcriptomic regulation, or “regulomes” based on the correlated snRNA– and snATAC-sequencing data. After calculating GPLs separately for each population (gray for RAGnaïve and tomato red for RAGexp), GPLs found in only one group, but not the other, can be identified (teal boxes). The difference in GPLs based on RAG experience for any given gene (e.g. Gene X) can be visualized on a bar graph, with the number of GPLs for RAGnaïve (gray – left) and RAGexp (red – right) plotted, with the difference overlaid as a black bar. C) GPLs calculated as in (B) for the multiomic ILC2 gene set identified in Figure 4H and Table S4). All identified GPLs are listed in Table S7, while ILC2 GPLs are listed in Table S8. Genes are sorted from more links identified in the RAGnaïve population at top to more links identified in the RAGexp population at bottom. Select genes are labeled. Full ranked list by difference in GPLs is available in Table S9.

We found that the ILC2 marker genes segregating toward the top of this list, corresponding to enhanced epigenomic activity in the RAGnaïve cells, tended to be genes previously identified to play positive roles in development, expansion, and activation of lymphoid cells. These included transcriptional regulators such as Tox75–78, Rora38,51,61–63, Maf19,58,59, and Gata319,50–52, which are involved in early differentiation of both ILCs and lymphocytes (Fig. S9A). Indeed, epigenetic activation of the Gata3 locus is recognized to play a critical role in development of both ILC2s79 and Th2 cells80,81. Additionally, surface receptors known to drive ILC2 activation upon stimulation including Il18r138,43,44, Il1rl121,38,43,82,83, and Icos39,40 had increased functional regulome activity in RAGnaïve ILC2s. In contrast, genes with more GPLs in the RAGexp ILC2s tended to be associated with suppressive functions. For example, Ndfip1 (Fig. S9B) encodes a regulatory protein that enhances activity of the ubiquitin ligase ITCH to negatively regulate inflammation84,85 and has been associated with asthma risk in GWAS studies86. Dusp1 partially mediates glucocorticoid effects through its ability to negatively regulate inflammation87,88, is associated with eczema by GWAS89, and has recently been shown to mark an anti-inflammatory set of ILCs69. Last, Asxl1 encodes a tumor suppressor that inhibits clonal hematopoiesis through its epigenomic regulatory effects in both mice and humans90–93. Collectively, our GPL analysis stratifies the ILC2 gene signature based on RAG experience, where genes associated with ILC2 expansion and activation are poised in RAGnaïve cells, while genes associated with suppressive effects are poised in RAGexp cells.

A history of RAG expression modulates ILC2 epigenomes at steady state and in AD-like inflammation

Although our GPL analysis revealed a suppressive effect of RAG expression on ILC2 gene programs, we did not account for the additional variable of disease state in the initial analysis. To test whether RAG expression promotes a suppressive epigenomic program in ILC2s that is durable in the setting of inflammation, we first recalculated GPLs after splitting our dataset by both history of RAG expression (naïve vs. experienced) and disease (steady state vs. AD-like disease) to yield four lists of GPLs (Fig 6A; Table S10). When we examined the intersection, or overlap, of peaks from ILC2 GPLs (Table S11), several notable patterns emerged (Fig. 6B). First, the largest set of peaks was shared by all RAGnaïve cells (gray bar), regardless of disease state, with the next largest peak sets belonging to either steady state or AD-like disease in the RAGnaïve cells. Second, there was a large set of peaks shared by all RAGexp cells (red bar). Third, the intersections corresponding to disease states (steady state – yellow, AD-like disease – dark red), had relatively few unique peaks. These findings suggest that early exposure to RAG expression plays a larger role in modulating the epigenomic signature of the ILC2 gene program than exposure to disease. To confirm that the patterns we observed represent a specific effect of RAG expression on the ILC2 gene program, we performed the same analysis on GPL peaks for all genes in the dataset (Fig. S10). In contrast to the ILC2 gene set, the majority of GPL peaks for all genes was shared among all cell populations, consistent with epigenomic regulation of most genes being minimally affected by either RAG expression or AD-like disease. Last, in the ILC2 gene set analysis, we noted a set of poised peaks shared by all RAGnaive cells and RAGexp cells in the setting of AD-like disease, but not with RAGexp cells at steady state (blue bar, Fig. 6B). We reasoned that this condition might capture some genomic loci that are suppressed by a history of RAG expression at steady state but are induced during inflammation.

A history of RAG expression broadly influences ILC2 genes at steady state and in AD-like inflammation.

A) Schematic of the process to determine contribution of RAG fate map and disease states to GPLs for subsequent intersection analyses. GPLs were first calculated for all indictated cells, regardless of disease state or fate map. Cells were then split, first by RAG fate map (RAGexp and RAGnaïve), and again by disease state (SS – steady state, AD – AD-like inflammation). GPLs were recalculated for each split sample and matched back to the original set of total GPLs. B) UpSet plot visualizing intersections of peaks identified from ILC2 GPLs for split samples. Each row represents one of the four sets, and each column corresponds to an intersection of one or more sets (see methods). See Table S10 for full list of GPLs for all genes. Table S11 lists total and ILC2 peaks used for intersection analyses in each of the four sets. Columns identifying key intersections are color coded by the corresponding RAG fate map or treatment groups. The blue column indicates the intersection of peaks from RAGnaïve cells and peaks induced by AD-like disease in RAGexp cells. C) Top genes with the most AD-like disease-induced peaks. Peaks from the intersection between RAGnaïve cells and inflamed RAGexp cells were identified in corresponding GPLs, and genes were ranked by number of linked peaks identified. See Table S12 for full list of ranked genes and associated GPLs. Open chromatin in the ILC2 cell cluster split by disease state and then by RAG fate map for the genomic loci of C) Ccr6 and D) Rora.

Thus, we next quantified and sorted these GPLs to generate a list of genes with the most peaks “induced” during AD-like disease (Fig. 6C, Table S12). Among the identified genes, we selected Rora (Fig. 6C) and Ccr6 (Fig. 6D) to examine more closely for evidence of epigenomic activation in AD-like disease, given the role of these genes in ILC2 expansion61,62 and homing to sites of inflammation44,94, respectively. For both genes, we observed more widespread open chromatin over the genomic region in the RAGnaïve cells compared to the RAGexp cells, but this difference was partially abolished by increased open chromatin in AD-like disease in the RAGexp cells. Taken together, our analysis reveals that a history of RAG expression selectively modulates activity of ILC2 gene programs across both steady state and during AD-like inflammation, while some programs are more evident at steady state given the uniquely poised nature of ILC2s.

RAG suppresses the Th2 locus

Our functional data demonstrate a role for RAG expression in regulating ILC2 development and activation, including limiting proportions of IL5+ and IL-13+ ILC2s at steady state and in AD-like disease. Prior work identified epigenomic priming in ILC2s early in development at the Th2 locus (comprised of the Il4, Il13, Rad50, and Il5 gene loci) to enable rapid transcriptional responses during inflammation30. Thus, we hypothesized that RAG promotes the functional observations in ILC2s by suppressing the establishment of an active regulome at the Th2 locus. To test this hypothesis, we analyzed the Th2 locus in our multiomic data in greater detail. Using a similar strategy to our analysis of ILC2 marker genes, we calculated the number of GPLs in the RAGexp and RAGnaïve cells, respectively, for the genes in the Th2 locus. (Fig. 7A, Table S7). We found many GPLs associated with the four Th2 locus genes, including significant crosstalk between these genes, similar to previous observations (Fig. 7A)95–97. Importantly, we identified fewer GPLs in the RAGexp cells, particularly for the Il5 and Il13 loci (Fig. 7A). As in our analysis of the ILC2 marker GPLs, we quantified the differences based on RAG fate mapping and found that all genes in this locus had increased GPLs in RAGnaïve cells relative to RAGexp cells (Fig. 6B; Table S13). Thus, our epigenomic data are consistent with a role for RAG expression in suppressing the type 2 regulome at the Th2 locus in ILC2s.

RAG suppresses the Th2 locus.

A) Coverage plot of the Th2 genomic locus. Open chromatin in the ILC2 cluster for each Rag1 fate-mapped state is shown on top, and corresponding peaks (teal) and gene to peaks links (GPLs) are shown below for the RAGnaïve sample (gray) and the RAGexp sample (tomato red). Only GPLs that fit in the coverage window are shown. B) All GPLs identified in each fate map state for the Th2 locus genes Il4, Il13, Rad50, and Il5. See Table S13 for full list of Th2 GPLs. The number of GPLs for each gene is shown on the left in gray for RAGnaïve and on the right in tomato red for RAGexp. The difference is superimposed in black, and genes are sorted from more GPLs identified in RAGnaïve at top to more links identified in RAGexp at bottom. C) UpSet plot of intersections of peaks identified from Th2 locus GPLs. GPLs were recalculated, this time for samples separately by both RAG fate map status (RAGexp and RAGnaïve) and disease (SS – steady state, AD – AD-like inflammation). Each row represents one of the four sets of peaks, and each column corresponds to an intersection of one or more sets (see methods). See Table S11 for full list of peaks from GPLs for all genes, including Th2 genes, in each of the four sets. Columns identifying key intersections are color coded by the corresponding RAG fate map or disease groups. The blue column indicates the intersection of peaks from RAGnaïve cells and peaks induced in AD-like disease in RAGexp cells. (D) Th2 genes sorted by number of AD-like disease-induced peaks. Peaks induced by AD-like disease were identified in corresponding GPLs, and genes were ranked by frequency of links to induced peaks (representation in identified GPLs). See Table S14 for full list of ranked Th2 locus genes and associated GPLs. E) Open chromatin tracks, split by disease (beige box – steady state; maroon box – AD-like disease) and by RAG fate map (RAGnaïve – gray, RAGexp – red) for Il13.

We next considered the additional effects of AD-like inflammation on the Th2 epigenomic regulome using the same approach we used to analyze the ILC2 gene set in Fig. 6. Again, we found the largest set of peaks was shared by the RAGnaïve cells, regardless of disease state, with the next largest peak sets belonging to either steady state or AD-like disease in the RAGnaïve cells (Fig. 7C). Furthermore, there was a large proportion of peaks shared by both RAGexp cells, consistent with a major contribution of a history of RAG expression to epigenomic modulation of the Th2 regulome. To quantify the potential effect of AD-like inflammation on reversing RAG-mediated suppression of Th2 locus genes, we mapped the 14 peaks shared by RAGnaive cells and RAGexp cells in the setting of AD-like inflammation (i.e. only suppressed in RAGnaive cells at steady state) back to the Th2 genes via their respective GPLs (Fig. 7C –blue bar, Table S14). Interestingly, Il13, which was not identified as a top ILC2 marker in our earlier analyses, had the highest number of linked peaks associated with potential induction in AD-like disease (Fig. 7D). When we examined the Il13 locus in the ILC2 cluster more closely, we found more widespread open chromatin in the RAGnaïve cells compared to the RAGexp cells (Fig. 7E). However, in the AD-like disease sample, the RAGexp cells displayed increased open chromatin relative to the steady state, consistent with induction in the setting of inflammation, like our earlier findings for ILC2 genes such as Ccr6 (Fig. 6E). Taken together, our functional data and multiomic analyses demonstrate a role for RAG expression in modulating genes critical for ILC2 development and function, including the key type 2 cytokines expressed from the Th2 locus.

Discussion

RAG recombinases evolved nearly 500 million years ago from endogenous transposons, crucially enabling antigen receptor rearrangement and emergence of the adaptive immune cell lineages present in all modern vertebrates29,98. Indeed, RAG deficiency leads to a complete lack of B and T lymphocytes, manifesting clinically as severe combined immunodeficiency (SCID)99–101. However, fate mapping studies have shown that multiple mature immune cell populations other than adaptive B and T lymphocytes have a history of RAG expression23,24,33,102,103. More recent studies by Karo et al found that RAG expression during NK cell development influences multiple cellular functions including antitumor cytotoxicity, cell proliferation, and survival24. Yet whether RAG modulates cellular functions of innate immune cell populations other than NK cells remains poorly understood. Here, using RAG-deficient mice, RAG fate mapping mice, and multiomic analyses, we report that RAG suppresses developmental and effector functions of ILC2s.

Our functional data in RAG-deficient mice demonstrate that populations of ILC2s producing the type 2 cytokines IL-5 and IL-13 preferentially expand in the absence of a history of RAG expression. This implies a specific role for RAG in developmental repression of ILC2s. Building on this, our multiomic RAG fate mapping analyses of ILC2 gene programs demonstrate extensive epigenomic differences between RAGexp and RAGnaïve cells. We found RAG-associated epigenomic suppression at multiple functional levels, including cell surface receptors, key transcription factors, and the Th2 locus encoding the type 2 cytokines Il5 and Il13. Although RAG is only transiently expressed early in lymphoid development104, our data demonstrate that RAG expression can imprint durable effects on ILC2 gene programs to restrain their function.

It is increasingly recognized that expression of effector molecules for both ILCs and their counterpart adaptive lymphocytes (e.g. IL-13 from ILC2s and Th2 cells) is governed by finely tuned transcriptomic and epigenomic regulomes30,105–107. ILCs tend to adopt these regulomes earlier in their development than T cells, and these “poised” regulatory elements are thought to underlie the ability of tissue-resident ILCs to rapidly respond to stimuli. In contrast, the regulomes of naïve T cells remain relatively inactive until stimulation. Given that T cells are uniformly RAG-experienced, our data provoke the hypothesis that RAGexp ILC2s adopt a phenotype closer to that of naive T cells and may require stronger stimuli than RAGnaïve ILC2s to become activated. Indeed, our analyses found the RAG-associated suppressive programs could be overcome in the setting of AD-like inflammation. Thus, sufficient RAG expression may mediate key events underlying the establishment and maintenance of functional regulomes not only in ILCs, but also T cells. How RAG might affect these changes, and whether they are independent of its enzymatic activity and/or antigen receptor recombination, remains to be elucidated.

Clinically, a link between enhanced type 2 immune activity and RAG dysfunction is well-established. Omenn Syndrome (OS) is a form of SCID characterized by exaggerated type 2 immune activation and typically arises in the setting of hypomorphic RAG gene mutations. Impaired antigen receptor rearrangement, with rare “leaky” recombination events, leads to expansion of autoreactive oligoclonal T cells, eosinophilia, and markedly elevated IgE108–112. Similar phenotypes have been observed in mice harboring RAG mutations analogous to those found in human patients with OS113,114. Notwithstanding these findings, the mechanisms underlying the propensity of oligoclonal T cells with hypomorphic RAG activity to preferentially develop into the Th2 subtype are unclear. Prior studies have found a role for regulatory T cells in controlling type 2 skewing of transferred T cells in RAG-deficient hosts, potentially explaining similar observations in patients with OS115. Our data provide an additional mechanism by which RAG dysfunction may lead to OS through loss of cell-intrinsic RAG-mediated suppression of type 2 cellular programs. Additionally, increased type 2 cytokine production from RAG-deficient ILC2s may, in trans, enhance expansion of the oligoclonal Th2 cell populations, IgE induction, and eosinophilia observed in RAG-deficient states like OS. However, whether other immune cell types with RAG dysfunction, such as ILCs, contribute to the pathogenesis of OS in humans has not been investigated.

Lymphoid acquisition of RAG activity may represent a newer evolutionary mechanism that fine tunes ancient innate immune cell programs in addition to enabling development of relatively newer antigen-specific adaptive immune cell populations. Independent of antigen receptor diversity, loss of this function may offer an explanation as to why oligoclonal T cells tend to expand and skew towards a Th2 cell phenotype in the setting of hypomorphic RAG function as in OS115. Further studies are needed to define whether the suppressive effects of RAG expression operate similarly in T and B cells. Although we demonstrate that this phenomenon is observed in ILC2s, whether hypomorphic RAG expression in bona fide Th2 cells not only results in oligoclonality but also loss of suppression of the Th2 locus independently of antigen receptor rearrangement remains an outstanding question. Indeed, during development of gene therapy strategies for RAG-deficient SCID, lower doses of wild type RAG transgene expression have been associated with development of OS-like conditions in transplanted RAG-deficient mice116–119.

A major limitation of our study is lack of a defined mechanism for how RAG expression imprints durable epigenomic and transcriptomic changes in ILC2s. The mechanisms by which RAG mediates VDJ recombination are well-defined, from the biochemical details of DNA-binding to the epigenomic accessibility of antigen receptor loci and timing of RAG expression29,120–122. Notwithstanding genomic stress24 or potential RAG dose effects116–119, how RAG expression might modulate broad developmental and functional lymphoid programs other than V(D)J recombination remains unclear. The RAG complex can bind both DNA and modified histones and has been observed to occupy thousands of sites across the genome27. Thus, RAG may directly influence open chromatin states or obscure transcription factor binding sites to alter ILC2 development and function. Interestingly, RAG preferentially binds near transcription start sites of open chromatin in mouse pre-B cells, although corresponding effects on gene expression were not observed27. Although canonical recombination sites are concentrated in the antigen receptor loci, cryptic recombination sites in other regions may be deleted or rearranged by RAG activity, altering transcriptional regulation of associated genes27. Finally, through its E3 ubiquitin ligase activity29, RAG may influence immune signaling pathways independently of transcription altogether. Beyond basic mechanisms of how RAG modulates ILC2 development and function, whether these effects may contribute to pathologic processes modulated by ILC2s in organs other than the skin remains unclear.

Our studies expand prior work implicating RAG in critical immune functions beyond antigen receptor rearrangement that is exclusive to adaptive lymphocytes. Further, we provide additional insights into why patients with OS exhibit atopic syndromes in the setting of adaptive lymphocyte deficiency. Future studies into mechanisms underlying these findings may lead to new therapeutic avenues for disorders such as atopic dermatitis, food allergy, and asthma.

eLife Assessment

This study provides new insights into the expression profile of ILCs that demonstrate a history of RAG expression. It examines in part the potential intrinsic regulation of RAG expression and seeks to understand how the epigenetic state of ILCs is established, although a full understanding of intrinsic factors is incomplete. The work provides an important molecular dataset, and with further strengthening of the understanding of intrinsic regulation, this paper would be of interest more broadly to cell biologists seeking to understand immune cell development.

Reviewer #1 (Public Review):

The study starts with the notion that in an AD-like disease model, ILC2s in the Rag1 knock-out were expanded and contained relatively more IL-5+ and IL-13+ ILC2s. This was confirmed in the Rag2 knock-out mouse model.

By using a chimeric mouse model in which wild-type knock-out splenocytes were injected into irradiated Rag1 knock-out mice, it was shown that even though the adaptive lymphocyte compartment was restored, there were increased AD-like symptoms and increased ILC2 expansion and activity. Moreover, in the reverse chimeric model, i.e. injecting a mix of wild-type and Rag1 knock-out splenocytes into irradiated wild-type animals, it was shown that the Rag1 knock-out ILC2s expanded more and were more active. Therefore, the authors could conclude that the RAG1 mediated effects were ILC2 cell-intrinsic.

Subsequent fate-mapping experiments using the Rag1Cre;reporter mouse model showed that there were indeed RAGnaïve and RAGexp ILC2 populations within naïve mice. Lastly, the authors performed multi-omic profiling, using single-cell RNA sequencing and ATAC-sequencing, in which a specific gene expression profile was associated with ILC2. These included well-known genes but the authors notably also found expression of Ccl1 and Ccr8 within the ILC2. The authors confirmed their earlier observations that in the RAGexp ILC2 population, the Th2 regulome was more suppressed, i.e. more closed, compared to the RAGnaïve population, indicative of the suppressive function of RAG on ILC2 activity. I do agree with the authors' notion that the main weakness was that this study lacks the mechanism by which RAG regulates these changes in ILC2s.

The manuscript is very well written and easy to follow, and the compelling conclusions are well supported by the data. The experiments are meticulously designed and presented. I wish to commend the authors for the study's quality.

Even though the study is compelling and well supported by the presented data, some additional context could increase the significance:

(1) The presence of the RAGnaïve and RAGexp ILC2 populations raises some questions on the (different?) origin of these populations. It is known that there are different waves of ILC2 origin (most notably shown in the Schneider et al Immunity 2019 publication, PMID 31128962). I believe it would be very interesting to further discuss or possibly show if there are different origins for these two ILC populations.

Several publications describe the presence and origin of ILC2s in/from the thymus (PMIDs 33432227 24155745). Could the authors discuss whether there might be a common origin for the RAGexp ILC2 and Th2 cells from a thymic lineage? If true that the two populations would be derived from different populations, e.g. being the embryonic (possibly RAGnaïve) vs. adult bone marrow/thymus (possibly RAGexp), this would show a unique functional difference between the embryonic derived ILC2 vs. adult ILC2.

(2) On line 104 & Figures 1C/G etc. the authors describe that in the RAG knock-out ILC2 are relatively more abundant in the lineage negative fraction. On line 108 they further briefly mentioned that this observation is an indication of enhanced ILC2 expansion. Since the study includes an extensive multi-omics analysis, could the authors discuss whether they have seen a correlation of RAG expression in ILC2 with regulation of genes associated with proliferation, which could explain this phenomenon?

Reviewer #2 (Public Review):

Summary:

The study by Ver Heul et al., investigates the consequences of RAG expression for type 2 innate lymphoid cell (ILC2) function. RAG expression is essential for the generation of the receptors expressed by B and T cells and their subsequent development. Innate lymphocytes, which arise from the same initial progenitor populations, are in part defined by their ability to develop in the absence of RAG expression. However, it has been described in multiple studies that a significant proportion of innate lymphocytes show a history of Rag expression. In compelling studies several years ago, members of this research team revealed that early Rag expression during the development of Natural Killer cells (Karo et al., Cell 2014), the first described innate lymphocyte, had functional consequences.

Here, the authors revisit this topic, a worthwhile endeavour given the broad history of Rag expression within all ILCs and the common use of RAG-deficient mice to specifically assess ILC function. Focusing on ILC2s and utilising state-of-the-art approaches, the authors sought to understand whether early expression of Rag during ILC2 development had consequences for activity, fitness, or function. Having identified cell-intrinsic effects in vivo, the authors investigated the causes of this, identifying epigenetic changes associated with the accessibility genes associated with core ILC2 functions.

The manuscript is well written and does an excellent job of supporting the reader through reasonably complex transcriptional and epigenetic analyses, with considerate use of explanatory diagrams. Overall I think that the conclusions are fair, the topic is thought-provoking, and the research is likely of broad immunological interest. I think that the extent of functional data and mechanistic insight is appropriate.

Strengths:

- The logical and stepwise use of mouse models to first demonstrate the impact on ILC2 function in vivo and a cell-intrinsic role. Initial analyses show enhanced cytokine production by ILC2 from RAG-deficient mice. Then through two different chimeric mice (including BM chimeras), the authors convincingly show this is cell intrinsic and not simply as a result of lymphopenia. This is important given other studies implicating enhanced ILC function in RAG-/- mice reflect altered competition for resources (e.g. cytokines).

- Use of Rag expression fate mapping to support analyses of how cells were impacted - this enables a robust platform supporting subsequent analyses of the consequences of Rag expression for ILC2.

- Use of snRNA-seq supports gene expression and chromatin accessibility studies - these reveal clear differences in the data sets consistent with altered ILC2 function.

- Convincing evidence of epigenetic changes associated with loci strongly linked to ILC2 function. This forms a detailed analysis that potentially helps explain some of the altered ILC2 functions observed in ex vivo stimulation assays.

- Provision of a wealth of expression data and bioinformatics analyses that can serve as valuable resources to the field.

Weaknesses:

- Lack of insight into precisely how early RAG expression mediates its effects, although I think this is beyond the scale of this current manuscript. Really this is the fundamental next question from the data provided here.

- The epigenetic analyses provide evidence of differences in the state of chromatin, but there is no data on what may be interacting or binding at these sites, impeding understanding of what this means mechanistically.

- Focus on ILC2 from skin-draining lymph nodes rather than the principal site of ILC2 activity itself (the skin). This may well reflect the ease at which cells can be isolated from different tissues.

- Comparison with ILC2 from other sites would have helped to substantiate findings and compensate for the reliance on data on ILC2 from skin-draining lymph nodes, which are not usually assessed amongst ILC2 populations.

- The studies of how ILC2 are impacted are a little limited, focused exclusively on IL-13 and IL-5 cytokine expression.

Reviewer #3 (Public Review):

In this study, Ver Heul et al. investigate the role of RAG expression in ILC2 functions. While RAG genes are not required for the development of ILCs, previous studies have reported a history of expression in these cells. The authors aim to determine the potential consequences of this expression in mature cells. They demonstrate that ILC2s from RAG1 or RAG2 deficient mice exhibit increased expression of IL-5 and IL-13 and suggest that these cells are expanded in the absence of RAG expression. However, it is unclear whether this effect is due to a direct impact of RAG genes or a consequence of the lack of T and B cells in this condition. This ambiguity represents a key issue with this study: distinguishing the direct effects of RAG genes from the indirect consequences of a lymphopenic environment.

The authors focus their study on ILC2s found in the skin-draining lymph nodes, omitting analysis of tissues where ILC2s are more enriched, such as the gut, lungs, and fat tissue. This approach is surprising given the goal of evaluating the role of RAG genes in ILC2s across different tissues. The study shows that ILC2s derived from RAG-/- mice are more activated than those from WT mice, and RAG-deficient mice show increased inflammation in an atopic dermatitis (AD)-like disease model. The authors use an elegant model to distinguish ILC2s with a history of RAG expression from those that never expressed RAG genes. However, this model is currently limited to transcriptional and epigenomic analyses, which suggest that RAG genes suppress the type 2 regulome at the Th2 locus in ILC2s.

The authors report a higher frequency of ILC2s in RAG-/- mice in skin-draining lymph nodes, which is expected as these mice lack T and B cells, leading to ILC expansion. Previous studies have reported hyper-activation of ILCs in RAG-deficient mice, suggesting that this is not necessarily an intrinsic phenomenon. For example, RAG-/- mice exhibit hyperphosphorylation of STAT3 in the gut, leading to hyperactivation of ILC3s. This study does not currently provide conclusive evidence of an intrinsic role of RAG genes in the hyperactivation of ILC2s. The splenocyte chimera model is artificial and does not reflect a normal environment in tissues other than the spleen. Similarly, the mixed BM model does not demonstrate an intrinsic role of RAG genes, as RAG1-/- BM cells cannot contribute to the B and T cell pool, leading to an expected expansion of ILC2s. As the data are currently presented it is expected that a proportion of IL-5-producing cells will come from the RAG1-/- BM.

Overall, the level of analysis could be improved. Total cell numbers are not presented, the response of other immune cells to IL-5 and IL-13 (except the eosinophils in the splenocyte chimera mice) is not analyzed, and the analysis is limited to skin-draining lymph nodes.

The authors have a promising model in which they can track ILC2s that have expressed RAG or not. They need to perform a comprehensive characterization of ILC2s in these mice, which develop in a normal environment with T and B cells. Approximately 50% of the ILC2s have a history of RAG expression. It would be valuable to know whether these cells differ from ILC2s that never expressed RAG, in terms of proliferation and expression of IL-5 and IL-13. These analyses should be conducted in different tissues, as ILC2s adapt their phenotype and transcriptional landscape to their environment. Additionally, the authors should perform their AD-like disease model in these mice.

The authors provide a valuable dataset of single-nuclei RNA sequencing (snRNA-seq) and ATAC sequencing (snATAC-seq) from RAGexp (RAG fate map-positive) and RAGnaïve (RAG fate map-negative) ILC2s. This elegant approach demonstrates that ILC2s with a history of RAG expression are epigenomically suppressed. However, key genes such as IL-5 and IL-13 do not appear to be differentially regulated between RAGexp and RAGnaïve ILC2s according to Table S5. Although the authors show that the regulome activity of IL-5 and IL-13 is decreased in RAGexp ILC2s, how do the authors explain that these genes are not differentially expressed between the RAGexp and RAGnaïve ILC2? I think that it is important to validate this in vivo.

Author response:

The following is the authors’ response to the original reviews

Public Reviews:

Reviewer #1 (Public Review):

The study starts with the notion that in an AD-like disease model, ILC2s in the Rag1 knockout were expanded and contained relatively more IL-5+ and IL-13+ ILC2s. This was confirmed in the Rag2 knock-out mouse model.

By using a chimeric mouse model in which wild-type knock-out splenocytes were injected into irradiated Rag1 knock-out mice, it was shown that even though the adaptive lymphocyte compartment was restored, there were increased AD-like symptoms and increased ILC2 expansion and activity. Moreover, in the reverse chimeric model, i.e. injecting a mix of wild-type and Rag1 knock-out splenocytes into irradiated wild-type animals, it was shown that the Rag1 knock-out ILC2s expanded more and were more active. Therefore, the authors could conclude that the RAG1 mediated effects were ILC2 cell-intrinsic.

Subsequent fate-mapping experiments using the Rag1Cre;reporter mouse model showed that there were indeed RAGnaïve and RAGexp ILC2 populations within naïve mice. Lastly, the authors performed multi-omic profiling, using single-cell RNA sequencing and ATACsequencing, in which a specific gene expression profile was associated with ILC2. These included well-known genes but the authors notably also found expression of Ccl1 and Ccr8 within the ILC2. The authors confirmed their earlier observations that in the RAGexp ILC2 population, the Th2 regulome was more suppressed, i.e. more closed, compared to the RAGnaïve population, indicative of the suppressive function of RAG on ILC2 activity. I do agree with the authors' notion that the main weakness was that this study lacks the mechanism by which RAG regulates these changes in ILC2s.

The manuscript is very well written and easy to follow, and the compelling conclusions are well supported by the data. The experiments are meticulously designed and presented. I wish to commend the authors for the study's quality.

Even though the study is compelling and well supported by the presented data, some additional context could increase the significance:

(1) The presence of the RAGnaïve and RAGexp ILC2 populations raises some questions on the (different?) origin of these populations. It is known that there are different waves of ILC2 origin (most notably shown in the Schneider et al Immunity 2019 publication, PMID 31128962). I believe it would be very interesting to further discuss or possibly show if there are different origins for these two ILC populations.

Several publications describe the presence and origin of ILC2s in/from the thymus (PMIDs 33432227 24155745). Could the authors discuss whether there might be a common origin for the RAGexp ILC2 and Th2 cells from a thymic lineage? If true that the two populations would be derived from different populations, e.g. being the embryonic (possibly RAGnaïve) vs. adult bone marrow/thymus (possibly RAGexp), this would show a unique functional difference between the embryonic derived ILC2 vs. adult ILC2.

We agree with the Reviewer that our findings raise important questions about ILC ontogeny. These are areas of ongoing investigation for us, and it is our hope this study may inform further investigation by others as well.

Regarding the Schneider et al study, we have considered the possibility that RAG expression may mark a particular wave of ILC2 origin. In that study, the authors used a tamoxifen-based inducible Cre strategy in their experiments to precisely time the lineage tracing of a reporter from the Rosa26 locus. Those lineage tracing mice would overlap genetically with the RAG lineage tracing mice we used in our current study, thus performing combined timed migration fate mapping and RAG fate mapping experiments would require creating novel mouse strains.

Similarly, the possible influence of the thymic or bone marrow environment on RAG expression in ILCs is an exciting possibility. Perhaps there are signals common to those environments that can influence all developing lymphocytes, including not only T and B cells but also ILCs, with one consequence being induction of RAG expression. While assessing levels of RAG-experienced ILCs in these tissues using our lineage tracing mouse may hint at these possibilities, conclusive evidence would require more precise control over the timing of RAG lineage tracing than our current reagents allow (e.g. to control for induction in those environments vs migration of previously fate-mapped cells to those environments).

To answer these questions directly, we are developing orthogonal lineage tracing mouse strains, which can report on both timing of ILC development and RAG expression, but these mice are not available yet. Given the limitations of our currently available reagents, we were careful to focus our manuscript on the skin phenotype and the more descriptive aspects of the RAG-induced phenotype. We have elaborated on these important questions and referenced all the studies noted by the Reviewer in the Discussion section as areas of future inquiry on lines 421-433.

(2) On line 104 & Figures 1C/G etc. the authors describe that in the RAG knock-out ILC2 are relatively more abundant in the lineage negative fraction. On line 108 they further briefly mentioned that this observation is an indication of enhanced ILC2 expansion. Since the study includes an extensive multi-omics analysis, could the authors discuss whether they have seen a correlation of RAG expression in ILC2 with regulation of genes associated with proliferation, which could explain this phenomenon?

We thank the Reviewer for pointing out this opportunity to further correlate our functional and multiomic findings. To address this, we first looked deeper into our prior analyses and found that among the pathways enriched in GSEA analysis of differentially expressed genes (DEGs) between RAG+ and RAG- ILC2s, one of the pathways suppressed in RAG+ ILC2s was “GOBP_EPITHELIAL_CELL_PROLIFERATION.”

( Author response image 1). There are a few other gene sets present in other databases such as MSigDB with terms including “proliferation,” but these are often highly specific to a particular cell type and experimental or disease condition (e.g. tissue-specific cancers). We did not find any of these enriched in our GSEA analysis.

GSEA plot of GOBP epithelial proliferation pathway in RAG-experienced vs RAG-naïve ILC2s.

The ability to predict cellular proliferation states from transcriptomic data is an area of active research, and there does not appear to be any universally accepted method to do this reliably. We found two recent studies (PMIDs 34762642; 36201535) that identified novel “proliferation signatures.” Since these gene sets are not present in any curated database, we repeated our GSEA analysis using a customized database with the addition of these gene sets. However, we did not find enrichment of these sets in our RAG+/- ILC2 DEG list. We also applied our GPL strategy integrating analysis of our epigenomic data to the proliferation signature genes, but we did not see any clear trend. Conversely, our GSEA analysis did not identify any enrichment for apoptotic signatures as a potential mechanism by which RAG may suppress ILC2s.

Notwithstanding the limitations of inferring ILC2 proliferation states from transcriptomic and epigenomic data, our experimental data suggest RAG exerts a suppressive effect on ILC2 proliferation. To formally test the hypothesis that RAG suppresses proliferation in the most rigorous way, we feel new mouse strains are needed that allow simultaneous RAG fate mapping and temporally restricted fate mapping. We elaborate on this in new additions to the discussion on lines 421-433.

Reviewer #2 (Public Review):

Summary:

The study by Ver Heul et al., investigates the consequences of RAG expression for type 2 innate lymphoid cell (ILC2) function. RAG expression is essential for the generation of the receptors expressed by B and T cells and their subsequent development. Innate lymphocytes, which arise from the same initial progenitor populations, are in part defined by their ability to develop in the absence of RAG expression. However, it has been described in multiple studies that a significant proportion of innate lymphocytes show a history of Rag expression. In compelling studies several years ago, members of this research team revealed that early Rag expression during the development of Natural Killer cells (Karo et al., Cell 2014), the first described innate lymphocyte, had functional consequences.

Here, the authors revisit this topic, a worthwhile endeavour given the broad history of Rag expression within all ILCs and the common use of RAG-deficient mice to specifically assess ILC function. Focusing on ILC2s and utilising state-of-the-art approaches, the authors sought to understand whether early expression of Rag during ILC2 development had consequences for activity, fitness, or function. Having identified cell-intrinsic effects in vivo, the authors investigated the causes of this, identifying epigenetic changes associated with the accessibility genes associated with core ILC2 functions.

The manuscript is well written and does an excellent job of supporting the reader through reasonably complex transcriptional and epigenetic analyses, with considerate use of explanatory diagrams. Overall I think that the conclusions are fair, the topic is thoughtprovoking, and the research is likely of broad immunological interest. I think that the extent of functional data and mechanistic insight is appropriate.

Strengths:

- The logical and stepwise use of mouse models to first demonstrate the impact on ILC2 function in vivo and a cell-intrinsic role. Initial analyses show enhanced cytokine production by ILC2 from RAG-deficient mice. Then through two different chimeric mice (including BM chimeras), the authors convincingly show this is cell intrinsic and not simply as a result of lymphopenia. This is important given other studies implicating enhanced ILC function in RAG-/- mice reflect altered competition for resources (e.g. cytokines).

- Use of Rag expression fate mapping to support analyses of how cells were impacted - this enables a robust platform supporting subsequent analyses of the consequences of Rag expression for ILC2.

- Use of snRNA-seq supports gene expression and chromatin accessibility studies - these reveal clear differences in the data sets consistent with altered ILC2 function.

- Convincing evidence of epigenetic changes associated with loci strongly linked to ILC2 function. This forms a detailed analysis that potentially helps explain some of the altered ILC2 functions observed in ex vivo stimulation assays.

- Provision of a wealth of expression data and bioinformatics analyses that can serve as valuable resources to the field.

We appreciate the strengths noted by the Reviewer for our study. We would like to especially highlight the last point about our single cell dataset and provision of supplemental data tables. Although our study is focused on AD-like skin disease and skin draining lymph nodes, we hope that our findings can serve as a valuable resource for future investigation into mechanisms of RAG modulation of ILC2s in other tissues and disease states.

Weaknesses:

- Lack of insight into precisely how early RAG expression mediates its effects, although I think this is beyond the scale of this current manuscript. Really this is the fundamental next question from the data provided here.

We thank the Reviewer for their recognition of the context of our current work and its future implications. We aimed to present compelling new observations within the scope of what our current data can substantiate. We believe answering the next fundamental question of the mechanisms by which RAG mediates its effects in ILC2s will require development of novel reagents. We are actively pursuing this, and we look forward to others building on our findings as well.

- The epigenetic analyses provide evidence of differences in the state of chromatin, but there is no data on what may be interacting or binding at these sites, impeding understanding of what this means mechanistically.

We thank the Reviewer for pointing out this aspect of the epigenomic data analysis and the opportunity to expand the scope of our manuscript. We performed additional analyses of our data to identify DNA binding motifs and infer potential transcription factors that may be driving the effects of a history of RAG expression that we observed. We hope that these additional data, analyses, and interpretation add meaningful insight for our readers.

We first performed the analysis for the entire dataset and validated that the analysis yielded results consistent with prior studies (e.g. finding EOMES binding motifs as a marker in NK cells). Then, we examined the differences in RAG fate-mapped ILC2s. These analyses are in new Figure S10 and discussed on lines 277-316.

We also performed an analysis specifically on the Th2 locus, given the effects of RAG on type 2 cytokine expression. These analyses are in new Figure S12 and discussed on lines 366-378.

- Focus on ILC2 from skin-draining lymph nodes rather than the principal site of ILC2 activity itself (the skin). This may well reflect the ease at which cells can be isolated from different tissues.

We appreciate the Reviewer’s insight into the limitations of our study. Difficulties in isolating ILC2s from the skin were indeed a constraint in our study. In particular, we were unable to isolate enough ILC2s from the skin for stimulation and cytokine staining. Given that one of our main hypotheses was that RAG affects ILC2 function, we focused our studies on skin draining lymph nodes, which allowed measurement of the two main ILC2 functional cytokines, IL-5 and IL-13, as readouts in the key steady state and AD-like disease experiments.

- Comparison with ILC2 from other sites would have helped to substantiate findings and compensate for the reliance on data on ILC2 from skin-draining lymph nodes, which are not usually assessed amongst ILC2 populations.

We agree with the Reviewer that a broader survey of the RAG-mediated phenotype in other tissues and by extension other disease models would strengthen the generalizability of our observations. Indeed, we did a more expansive survey of tissues in our BM chimera experiments. We found a similar trend to our reported findings in the sdLN in tissues known to be affected by ILC2s ( Author response image 2) including the skin and lung and in other lymphoid tissues including spleen and mesenteric lymph nodes (mLN). We found that donor reconstitution in each tissue was robust except for the skin, where there was no significant difference between host and -donor CD45+ immune cells and where CD45- parenchymal cells predominated ( Author response image 2A,C,E,G,I). This may explain why Rag1-/- donor ILC2s were significantly higher in proportion in all tissues except the skin, where we observed a similar trend that was not statistically significant ( Author response image 2B,D,F,H,J).

Notwithstanding these results, given that we unexpectedly observed enhanced AD-like inflammation in the MC903 model in Rag1 KO mice, we concentrated our later experiments and analyses on defining the differences in skin draining ILC2s modulated by RAG. Our subsequent findings in the skin provoke many new hypotheses about the role of RAG in ILC2s in other tissues, and our tissue survey in the BM chimera provides additional rationale to pursue similar studies in disease models in other tissues. While this is an emerging area of investigation in our lab, we opted to focus this manuscript on our findings related to the AD-like disease model. We have ongoing studies to investigate other tissues, and we are still in the early stages of developing disease models to expand on these findings. However, if the reviewer feels strongly this additional data should be included in the manuscript, we are happy to add it. Considering the complexity of the data and concepts in the manuscript, we hoped to keep it focused to where we have strong molecular, cellular, and phenotypic outcomes.

Comparison of immune reconstitution in and ILC2 donor proportions in different tissues from BM chimeras.

Equal quantities of bone marrow cells from Rag1-/- (CD45.2,CD90.2) and WT (CD45.2, CD90.1) C57Bl/6J donor mice were used to reconstitute the immune systems of irradiated recipient WT (CD45.1) C57Bl/6J mice. The proportion of live cells that are donor-derived (CD45.2), host-derived (CD45.1), or parenchymal (CD45-) [above] and proportion of ILC2s that are from Rag1-/- (CD90.2) or WT (CD90.1) donors [below] for A,B) skin C,D) sdLN E,F) lung G,H) spleen and I,J) mLN.

- The studies of how ILC2 are impacted are a little limited, focused exclusively on IL-13 and IL-5 cytokine expression.

We agree with the reviewer that our functional readout on IL-5 and IL-13 is relatively narrow. However, this focused experimental design was based on several considerations. First, IL-5 and IL-13 are widely recognized as major ILC2 effector molecules (Vivier et al, 2018, PMID 30142344). Second, in the MC903 model of AD-like disease, we have previously shown a clear correlation between ILC2s, levels of IL-5 and IL-13, and disease severity as measured by ear thickness (Kim et al, 2013, PMID 23363980). Depletion of ILC2s led to decreased levels of IL-13 and IL-5 and correspondingly reduced ear inflammation. However, while ILC2s are also recognized to produce other effector molecules such as IL-9 and Amphiregulin, which are likely involved in human atopic dermatitis (Namkung et al, 2011, PMID 21371865; Rojahn et al, 2020, PMID 32344053), there is currently no evidence linking these effectors to disease severity in the MC903 model. Third, IL-13 is emerging as a key cytokine driving atopic dermatitis in humans (Tsoi et al, 2019, PMID 30641038). Drugs targeting the IL-4/IL-13 receptor (dupilumab), or IL-13 itself (tralokinumab, lebrikizumab), have shown clear efficacy in treating atopic dermatitis. Interestingly, drugs targeting more upstream molecules, like TSLP (tezepelumab) or IL-33 (etokimab), have failed in atopic dermatitis. Taken together, these findings from both mouse and human studies suggest IL-13 is a critical therapeutic target, and thus functional readout, in determining the clinical implications of type 2 immune activation in atopic dermatitis.

Aside from effector molecules, other readouts such as surface receptors may be of interest in understanding the mechanism of how RAG influences ILC2 function. For example, IL-18 has been shown to be an important co-stimulatory molecule along with TSLP in driving production of IL-13 by cutaneous ILC2s (Ricardo-Gonzalez et al, 2018, PMID 30201992). Our multiomic analysis showed decreased IL-18 receptor regulome activity in RAG-experienced ILC2s, which may be a mechanism by which RAG suppresses IL-13 production. Ultimately, in that study the role of IL-18 in enhancing MC903-induced inflammation through ILC2s was via increased production of IL-13, which was one of our major functional readouts. To clearly define mechanisms like these will require generation of new mice to interrogate RAG status in the context of tissue-specific knockout of other genes, such as the IL-18 receptor. We plan to perform these types of experiments in follow up studies. Notwithstanding this, we have now included additional discussion on lines 476508 to highlight why understanding how RAG impacts other regulatory and effector pathways would be an interesting area of future inquiry.

Reviewer #3 (Public Review):

In this study, Ver Heul et al. investigate the role of RAG expression in ILC2 functions. While RAG genes are not required for the development of ILCs, previous studies have reported a history of expression in these cells. The authors aim to determine the potential consequences of this expression in mature cells. They demonstrate that ILC2s from RAG1 or RAG2 deficient mice exhibit increased expression of IL-5 and IL-13 and suggest that these cells are expanded in the absence of RAG expression. However, it is unclear whether this effect is due to a direct impact of RAG genes or a consequence of the lack of T and B cells in this condition. This ambiguity represents a key issue with this study: distinguishing the direct effects of RAG genes from the indirect consequences of a lymphopenic environment.

The authors focus their study on ILC2s found in the skin-draining lymph nodes, omitting analysis of tissues where ILC2s are more enriched, such as the gut, lungs, and fat tissue. This approach is surprising given the goal of evaluating the role of RAG genes in ILC2s across different tissues. The study shows that ILC2s derived from RAG-/- mice are more activated than those from WT mice, and RAG-deficient mice show increased inflammation in an atopic dermatitis (AD)-like disease model. The authors use an elegant model to distinguish ILC2s with a history of RAG expression from those that never expressed RAG genes. However, this model is currently limited to transcriptional and epigenomic analyses, which suggest that RAG genes suppress the type 2 regulome at the Th2 locus in ILC2s.

We agree with the Reviewer that understanding the role of RAG in ILC2s across different tissues is an important goal. One of the primary inspirations for our paper was the clinical paradox that patients with Omenn syndrome, despite having profound adaptive T cell deficiency, develop AD with much greater penetrance than in the general population. Thus, there was always an appreciation for the likelihood that skin ILC2s have a unique proclivity towards the development of AD-like disease. Notwithstanding this, given the profound differences that can be found in ILC2s based on their tissue residence and disease state (as the Reviewer also points out below), we focused our investigations on characterizing the skin draining lymph nodes to better define factors underlying our initial observations of enhanced AD-like disease in Rag1-/- mice. While our findings in skin provoke the hypothesis that similar effects may be observed in other tissues and influence corresponding disease states, we were cautious not to suggest this may be the case by reporting surveys of other tissues without development of additional disease models to formally test these hypotheses. We present this manuscript now as a short, skin-focused study, rather than delaying publication to expand its scope. Truthfully, this project started in 2015 and has undergone many delays with the hopes of newer technologies and reagents coming to add greater clarity. We hope our study will enable others to pursue the goal of understanding the broader effects of RAG in ILC2s, and potentially other innate lymphoid lineages as well.

We did a more expansive survey of tissues in our BM chimera experiments. We found a similar trend to our reported findings in the sdLN in tissues known to be affected by ILC2s ( Author response image 2) including the skin and lung and in other lymphoid tissues including spleen and mesenteric lymph nodes (mLN). We found that donor reconstitution in each tissue was robust except for the skin, where there was no significant difference between host and donor CD45+ immune cells and where CD45- parenchymal cells predominated ( Author response image 2A,C,E,G,I). This may explain why Rag1-/- donor ILC2s were significantly higher in proportion in all tissues except the skin, where we observed a similar trend that was not statistically significant ( Author response image 2B,D,F,H,J). However, given the lack of correlation to disease readouts in other organ systems, we chose to not include this data in our manuscript. However, if the Reviewer feels these data should be included, we would be happy to include as a supplemental figure.

The authors report a higher frequency of ILC2s in RAG-/- mice in skin-draining lymph nodes, which is expected as these mice lack T and B cells, leading to ILC expansion. Previous studies have reported hyper-activation of ILCs in RAG-deficient mice, suggesting that this is not necessarily an intrinsic phenomenon. For example, RAG-/- mice exhibit hyperphosphorylation of STAT3 in the gut, leading to hyperactivation of ILC3s. This study does not currently provide conclusive evidence of an intrinsic role of RAG genes in the hyperactivation of ILC2s. The splenocyte chimera model is artificial and does not reflect a normal environment in tissues other than the spleen. Similarly, the mixed BM model does not demonstrate an intrinsic role of RAG genes, as RAG1-/- BM cells cannot contribute to the B and T cell pool, leading to an expected expansion of ILC2s. As the data are currently presented it is expected that a proportion of IL-5-producing cells will come from the RAG1/- BM.

The Reviewer raises an important point about the potential cell-intrinsic roles of RAG vs the many cell-extrinsic explanations that could affect ILC2 populations, with the most striking being the lack of T and B cells in RAG knockout mice. It is well-established that splenocyte transfer into T and B cell-deficient mice reconstitutes T cell-mediated effects (such as the T cell transfer colitis model pioneered by Powrie and others), and we were careful in our interpretation of the splenocyte chimera experiment to conclude only that lack of Tregs was unlikely to explain the enhanced ADlike disease in T (and B) cell-deficient mice.

We agree with the Reviewer that the Rag1-/- BM will not contribute to the B and T cell pool. However, BM from the WT mice would be expected to contribute to development of the adaptive lymphocyte pool. Indeed, we found that most of the CD45+ immune cells in the spleens of BM chimera mice were donor-derived ( Author response image 3A), and total levels of B cells and T cells showed reconstitution in a pattern similar to control spleens from donor WT mice, while spleens from donor Rag1-/- mice expectedly had essentially no detectable adaptive lymphocytes ( Author response image 3B-D). From this, we concluded the BM chimera experiment was successful in establishing an immune environment with the presence of adaptive lymphocytes, and the differences in ILC2 proportions we observed were in the context of developing alongside a normal number of B and T lymphocytes. Notwithstanding the potential role of the adaptive lymphocyte compartment in shaping ILC2 development, since we transplanted equal amounts of WT and Rag1-/- BM into the same recipient environment, we are not able to explain how cell-extrinsic effects alone would account for the unequal numbers of WT vs Rag1-/- ILC2s we observed after immune reconstitution.

Comparison of immune reconstitution in BM chimeras to controls.

Equal quantities of bone marrow cells from Rag1-/- (CD45.2) and WT (CD45.2) C57Bl/6J donor mice were used to reconstitute the immune systems of irradiated recipient WT (CD45.1) C57Bl/6J mice. A) Number of WT recipient CD45.1+ immune cells in the spleens of recipient mice compared to number of donor CD45.2+ cells (WT and Rag1-/-) normalized to 100,000 live cells. Comparison of numbers of B cells, CD4+ T cells, and CD8+ T cells in spleens of B) BM chimera mice, C) control WT mice and D) control Rag1-/- mice.

We also subsequently found transcriptional and epigenomic differences in RAG-experienced ILC2s compared to RAG-naïve ILC2s. Critically, these differences were present in ILC2s from the same mice that had developed normally within an intact immune system, rather than in the setting of a BM transplant or a defective immune background such as in Rag1-/- mice.

We recognize that there are almost certainly cell-extrinsic factors affecting ILC2s in Rag1-/- mice due to lack of B and T cells, and that BM chimeras are not perfect substitutes for simulating normal hematopoietic development. However, the presence of cell-extrinsic effects does not negate the potential contribution of cell-intrinsic factors as well, and we respectfully stand by our conclusion that our data support a role, however significant, for cell-intrinsic effects of RAG in ILC2s.

Finally, the Reviewer mentions the interesting observation that gut ILC3s exhibit hyperphosphorylation of STAT3 in Rag1-/- mice compared to WT as an example of cell-extrinsic effects of RAG deficiency (we assume this is in reference to Mao et al, 2018, PMID 29364878 and subsequent work). We now reference this paper and have included additional discussion on how our observations of ILC2s may be generalizable to not only other organ systems, but also other ILC subsets, limitations on these generalizations, and future directions on lines 477-520.

Overall, the level of analysis could be improved. Total cell numbers are not presented, the response of other immune cells to IL-5 and IL-13 (except the eosinophils in the splenocyte chimera mice) is not analyzed, and the analysis is limited to skin-draining lymph nodes.

We thank the Reviewer for the suggestions to add rigor to our analysis. ILC2 populations are relatively rare, and we designed our experiments to assess frequencies, rather than absolute numbers. We did not utilize counting beads, so our counts may not be comparable between samples. We have added additional data for absolute cell counts normalized to 100,000 live cells for each experiment (see below for a summary of new panels in each figure). Our new data on total cell numbers are consistent with the initial observations regarding frequency of ILC2s we reported from our experiments. For the BM chimera experiments, we presented the proportions of ILC2s, and IL-5 and IL-13 positive ILC2s, by donor source, as this is the critical question of the experiment. Notwithstanding our analysis by proportion, we found that the frequency of Rag1-/- ILC2s, IL-5+ cells, or IL-13+ cells within Lin- population was also significantly increased. While our initial submission included only the proportions for clarity and simplicity, we now include frequency and absolute numbers in new panels for more critical appraisal of our data by readers.

In New Figure 1, we added new panels for ILC2 cell number in both the AD-like disease experiment (C) and in steady state (H).

In New Figure S2, we added a panel for ILC2 cell number in steady state (B).

In Figure 2 and associated supplemental data in Figure S4, we added several more panels. For the splenocyte chimera, we added a panel for ILC2 cell number in New Figure 2C.

We incorporated multiple new panels in New Figure S4 to address the need for more data to be shown for the BM chimera (also requested by Reviewer #2). These included total cell counts and frequency for ILC2 (New Figure S4F,G), and IL-5+ (New Figure S4I,K) and IL-13+ (New Figure S4J,L) ILCs in addition to the proportions originally presented in Figure 2.

In terms of the limited analysis of other tissues, our initial observation of enhanced AD-like disease in Rag1-/- compared to WT mice built on our prior work elucidating the role of ILC2s in the MC903 model of AD-like disease in mice and AD in humans (Kim et al, 2013, PMID 23363980). Consequently, we focused on the skin to further develop our understanding of the role of RAG1 in this model. As in our prior studies, technical limitations in obtaining sufficient numbers of ILC2s from the skin itself for ex vivo stimulation to assess effector cytokine levels required performing these experiments in the skin draining lymph nodes.

We agree that IL-5 and IL-13 are major mediators of type 2 pathology and studying their effects on immune cells is an important area of inquiry, particularly since there are multiple drugs available or in development targeting these pathways. However, our goal was not to study what was happening downstream of increased cytokine production from ILC2s, but instead to understand what was different about RAG-deficient or RAG-naïve ILC2s themselves that drive their expansion and production of effector cytokines compared to RAG-sufficient or RAGexperienced ILC2s. By utilizing the same MC903 model in which we previously showed a critical role for ILC2s in driving IL-5 and IL-13 production and subsequent inflammation in the skin, we were able to instead focus on defining the cell-intrinsic aspects of RAG function in ILC2s.

The authors have a promising model in which they can track ILC2s that have expressed RAG or not. They need to perform a comprehensive characterization of ILC2s in these mice, which develop in a normal environment with T and B cells. Approximately 50% of the ILC2s have a history of RAG expression. It would be valuable to know whether these cells differ from ILC2s that never expressed RAG, in terms of proliferation and expression of IL5 and IL-13. These analyses should be conducted in different tissues, as ILC2s adapt their phenotype and transcriptional landscape to their environment. Additionally, the authors should perform their AD-like disease model in these mice.

We agree with the Reviewer (and a similar comment from Reviewer #2) that a broader survey of the RAG-mediated phenotype in other tissues and by extension other disease models would strengthen the generalizability of our observations. Indeed, we did a more expansive survey of tissues in our BM chimera experiments. We found a similar trend to our reported findings in the sdLN in tissues known to be affected by ILC2s ( Author response image 2) including the skin and lung and in other lymphoid tissues including spleen and mesenteric lymph nodes (mLN). We found that donor reconstitution in each tissue was robust except for the skin, where there was no significant difference between host and donor CD45+ immune cells and where CD45- parenchymal cells predominated (Author response image 2A,C,E,G,I). This may explain why Rag1-/- donor ILC2s were significantly higher in proportion in all tissues except the skin, where we observed a similar trend that was not statistically significant (Author response image 2B,D,F,H,J). We omitted these analyses to maintain the focus on the skin, but we will be happy to add this data to the manuscript if the Reviewer feels this figure should be helpful.

Notwithstanding these results, given that we unexpectedly observed enhanced AD-like inflammation in the MC903 model in Rag1 KO mice, we concentrated our later experiments and analyses on defining the differences in skin draining ILC2s modulated by RAG. Our subsequent findings in the skin provoke many new hypotheses about the role of RAG in ILC2s in other tissues, and our tissue survey in the BM chimera provides additional rationale to pursue similar studies in disease models in other tissues. While this is an emerging area of investigation in our lab, we opted to focus this manuscript on our findings related to the AD-like disease model. We have ongoing studies to investigate other tissues, and we are still in the early stages of developing disease models to expand on these findings. However, if the reviewer feels strongly this additional data should be included in the manuscript, we are happy to add it. Considering the complexity of the data and concepts in the manuscript, we hoped to keep it focused to where we have strong molecular, cellular, and phenotypic outcomes. We elaborate on the implications of our work for future studies, including limitations of our study and currently available reagents and need for new mouse strains to rigorously answer these questions on lines 476-508

The authors provide a valuable dataset of single-nuclei RNA sequencing (snRNA-seq) and ATAC sequencing (snATAC-seq) from RAGexp (RAG fate map-positive) and RAGnaïve (RAG fate map-negative) ILC2s. This elegant approach demonstrates that ILC2s with a history of RAG expression are epigenomically suppressed. However, key genes such as IL-5 and IL-13 do not appear to be differentially regulated between RAGexp and RAGnaïve ILC2s according to Table S5. Although the authors show that the regulome activity of IL-5 and IL-13 is decreased in RAGexp ILC2s, how do the authors explain that these genes are not differentially expressed between the RAGexp and RAGnaïve ILC2? I think that it is important to validate this in vivo.

We thank the Reviewer for highlighting the value and possible elegance of our data. The Reviewer brings up an important issue that we grappled with in this study and that highlights a major technical limitation of single cell sequencing studies. Genes for secreted factors such as cytokines are often transcribed at low levels and are poorly detected in transcriptomic studies. This is particularly true in single cell studies with lower sequencing depth. Various efforts have been made to overcome these issues such as computational approaches to estimate missing data (e.g. van Djik et al, 2018, PMID 29961576; Huang et al, 2018, PMID 29941873), or recent use of cytokine reporter mice and dial-out PCR to enhance key cytokine signals in sequenced ILCs (Bielecki et al, 2021, PMID 33536623). We did not utilize computational methods to avoid the risk of introducing artifacts into the data, and we did not perform our study in cytokine reporter mice. Thus, cytokines were poorly detected in our transcriptomic data, as evidenced by lack of identification of cytokines as markers for specific clusters (e.g. IL-5 for ILC2s) or significant differential expression between RAG-naïve and RAG-experienced ILC2s.

However, the multiomic features of our data allowed a synergistic analysis to identify effects on cytokines. For example, transcripts for the IL-4 and IL-5 were not detected at a high enough level to qualify as marker genes of the ILC2 cluster in the gene expression (GEX) assay but were identified as markers for the ILC2 cluster in the ATAC-seq data in the differentially accessible chromatin (DA) assay. Using the combined RNA-seq and ATAC-seq gene to peak links (GPL) analyses, many GPLs were identified in the Th2 locus for ILC2s, including for IL-13, which was not identified as a marker for ILC2s by any of the assays alone. Thus, our combined analysis took advantage of the potential of multiomic datasets to overcome a general weakness inherent to most scRNAseq datasets.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

- Line 168; Reference 23 also showed expression in the NK cells, please add this reference to reference 24.

We thank the reviewer for catching this oversight, and we have corrected it in the revised manuscript.

- Please add the full names for GPL and sdLN in the text of the manuscript when first using these abbreviations. They are now only explained in the legends.

We reviewed the manuscript text and found that we defined sdLNs for the first time on line 104. We defined GPLs for the first time on line 248. We believe these definitions are placed appropriately near the first references to the corresponding figures/analysis, but if the Reviewer believes we should move these definitions earlier, we are happy to do so.

Reviewer #2 (Recommendations For The Authors):

I would suggest that the following reanalyses would improve the clarity of the data:

- Can ILC2 numbers, rather than frequency, be used (e.g. in Figure 1C, S2B, and so on). This would substantiate the data that currently relies on percentages.

This was a weakness also noted by Reviewer #3. We have added data on ILC2 numbers for each experiment as outlined below:

In New Figure 1, we added new panels for ILC2 cell number in both the AD-like disease experiment (C) and in steady state (H).

In New Figure S2, we added a panel for ILC2 cell number in steady state (B).

In Figure 2 and associated supplemental data in Figure S4, we added several more panels. For the splenocyte chimera, we added a panel for ILC2 cell number in New Figure 2C.

We incorporated multiple new panels in New Figure S4 to address the need for more data to be shown for the BM chimera (also requested by Reviewer #2). These included total cell counts and frequency for ILC2 (New Figure S4F,G), and IL-5+ (New Figure S4I,K) and IL-13+ (New Figure S4J,L) ILCs in addition to the proportions originally presented in Figure 2.

- Can the authors provide data on IL-33R expression on sdLN ILC2s? Expression of ST-2 (IL-33R) does vary between ILC2 populations and is impacted by the digestion of tissue. All of the data provided here requires ILC2 to be IL-33R+. In the control samples, the ILC2 compartment is very scarce - in LNs, ILC2s are rare. The gating strategy with limited resolution of positive and negative cells in the lineage gate doesn't help this analysis.

The Reviewer raises a valid point regarding the IL-33R marker and ILC2s. We designed our initial experiments to be consistent with our earlier observations of skin ILC2s, which were defined as CD45+Lin-CD90+CD25+IL33+, and the scarcity of skin draining lymph node ILC2s at steady state was consistent with our prior findings (Kim et al, 2013, PMID 23363980). We can include MFI data on IL-33R expression in these cells if the reviewer feels strongly that this would add to the manuscript, but we did not include other ILC2-specific markers in these experiments that would give us an alternative total ILC2 count to calculate frequency of IL-33R+ ILC2s, which would also make the context of the IL-33 MFI difficult to interpret.

Other studies defining tissue specific expression patterns in ILC2s have called into question whether IL-33R is a reliable marker to define skin ILC2s (Ricardo-Gonzalez et al, 2018, PMID 30201992). However, there is evidence for region-specific expression of IL-33R (Kobayashi et al, 2019, PMID 30712873), with ILC2s in the subcutis expressing high levels of IL-33R and both IL5 and IL-13, while ILC2s in the epidermis and dermis have low levels of IL-33R and IL-5 expression. In contrast to the Kobayashi et al study, Ricardo-Gonzalez et al sequenced ILC2s from whole skin, thus the region-specific expression patterns were not preserved, and the lower expression of IL-33R in the epidermis and dermis may have diluted the signal from the ILC2s in the subcutis. These may also be the ILC2s most likely to drain into the lymph nodes, which is the tissue on which we focused our analyses (consistent with our prior work in Kim et al, 2013).

- In Figure 2 (related to 2H, 2I) can flow plots of the IL-5 versus IL-13 gated on either CD90.1+CD45.2+ or CD90.2+CD45.2+ ILC2 be shown? I.e. gate on the ILC2s and show cytokine expression, rather than the proportion of donor IL5/13. The proportion of donor ILC2 is shown to be significantly higher in 2G. Therefore gating on the cells of interest and showing on a cellular basis their ability to produce the cytokines would better make the point I think.

We agree that this is important additional data to include. We have added flow plots of sdLN ILC2s from the BM chimera divided by donor genotype showing IL-5 and IL-13 expression in New Figure S4H.

I assume the authors have looked and there is no obvious data, but does analysis of transcription factor consensus binding sequences in the open chromatin provide any new insight?

The Reviewer also commented on this in the public review. As copied from our response above:

We found that the most enriched sites in the ILC2 gene loci contained the consensus sequence GGGCGG (or its reverse complement), a motif recognized by a variety of zinc finger transcription factors (TFs). Predictions from our analyses predicted the KLF family of zinc finger TFs as most likely to be enriched at the identified open chromatin regions. To infer which KLFs might be occupying these sites in the RAG-experienced or RAG-naïve cells, we also assessed the expression levels of these identified TFs. Interestingly, KLF2 and KLF6 are more expressed in RAG-experienced ILC2s. KLF6 is a tumor suppressor (PMID: 11752579), and both KLF6 and KLF2 were recently shown to be markers of “quiescent-like” ILCs (PMID: 33536623). Further, upon analysis of the Th2 locus, the (A/T)GATA(A/G) consensus site (or reverse complement) was enriched in identified open chromatin at that locus. The algorithm predicted multiple TFs from the GATA family as possible binding partners, but expression analysis showed only GATA3 was highly expressed in ILC2s, consistent with what would be predicted from prior studies (PMID: 9160750).

We have added this data in new Figure S10 and new Figure S12, with corresponding text in the Results section on lines 277-316 and lines 366-378.

In terms of phrasing and presentation:

- It would help to provide some explanation of why all analyses focus on the draining LNs rather than the actual site of inflammation (the ear skin). I do not think it appropriate to ask for data on this as this would require extensive further experimentation, but there should be some discussion on this topic. This feels relevant given that the skin is the site of inflammatory insult and ILC2 is present here. How the ILC2 compartment in the skindraining lymph nodes relates to those in the skin is not completely clear, particularly given the prevailing dogma that ILC2 are tissue-resident.

Given limitations of assessing cytokine production of the relatively rare population of skin-resident ILC2s, we focused on the skin-draining lymph nodes (sdLN). Our findings in the current manuscript are consistent with our prior work in Kim et al, 2013 (PMID 23363980), and more recently in Tamari et al, 2024 (PMID 38134932), which demonstrated correlation of increased ILC2s in sdLN with increased skin inflammation in the MC903 model. Similarly, Dutton et al (PMID 31152090) have demonstrated expansion of the sdLN ILC2 pool in response to MC903-induced AD-like inflammation in mice. We elaborate on the implications of our work for future studies, including limitations of our study (including the focus on the sdLN), and currently available reagents and need for new mouse strains to rigorously answer these questions on lines 476-508

- I think the authors should explicitly state that cytokine production is assessed after ex vivo restimulation (e.g. Lines 112-113).

We have added this statement to the revised text.

- I also think that it would help to be consistent with axis scales where analyses are comparable (e.g. Figure 1D vs Figure 1H).

We agree with the Reviewer and we have adjusted the axes for consistency. The data remains unchanged, but axes are slightly adjusted in New Figure 1 (D&I, E&J, F&K) and New Figure S2 (C-E match New Figure 1 D-F). This same axis scaling scheme is carried forward to New Figure 2 (D-E) and New Figure S4 (G,K,L). New data on cell counts is also included per request by Reviewers 2 and 3 (see above). However, we found results for total cells, including ILC2s (New Figure 1C,H, New Figure S2B, New Figure 2C, New Figure S4F), were consistent within experiments, but not between experiments, likely representing issues with normalizing counts (we did not include counting beads for more accurate total counts). Thus, the y-axes in those panels are not consistent between experiments/figures.

We feel reporting the proportion of WT vs Rag1-/- donor cells for the BM chimera is most illustrative of the effect of RAG and have kept it in the main New Figure 2, but for the BM chimera experiment panels we also include the total counts of IL-5+ and IL-13+ ILC2s (New Figure S4I,J).

Version 2

Abstract

Abstract

Antigen specificity is the central trait distinguishing adaptive from innate immune function. Assembly of antigen-specific T cell and B cell receptors occurs through V(D)J recombination mediated by the Recombinase Activating Gene endonucleases RAG1 and RAG2 (collectively called RAG). In the absence of RAG, mature T and B cells do not develop and thus RAG is critically associated with adaptive immune function. In addition to adaptive T helper 2 (Th2) cells, group 2 innate lymphoid cells (ILC2s) contribute to type 2 immune responses by producing cytokines like Interleukin-5 (IL-5) and IL-13. Although it has been reported that RAG expression modulates the function of innate natural killer (NK) cells, whether other innate immune cells such as ILC2s are affected by RAG remains unclear. We find that in RAG-deficient mice, ILC2 populations expand and produce increased IL-5 and IL-13 at steady state and contribute to increased inflammation in atopic dermatitis (AD)-like disease. Further, we show that RAG modulates ILC2 function in a cell-intrinsic manner independent of the absence or presence of adaptive T and B lymphocytes. Lastly, employing multiomic single cell analyses of RAG1 lineage-traced cells, we identify key transcriptional and epigenomic ILC2 functional programs that are suppressed by a history of RAG expression. Collectively, our data reveal a novel role for RAG in modulating innate type 2 immunity through suppression of ILC2s.

Introduction

Atopic disorders such as atopic dermatitis (AD), asthma, and food allergy are associated with T helper type 2 (Th2) cell responses, elevated production of the type 2 cytokines interleukin(IL)-4, IL-5, and IL-13, and induction of immunoglobulin(Ig)E1–4.Classically, this allergic inflammatory cascade is believed to originate with antigenic stimulation of T cell receptors on adaptive T cells, which in turn results in the production of IgE from B and plasma cells capable of binding the same antigen. Indeed, presence of antigen-specific IgE reactivity is a hallmark across atopic disorders.5 Thus, for decades, antigen-specific adaptive Th2 cell responses have been the primary focus of investigation in the pathogenesis of atopic diseases. However, recent studies indicate that innate immune cells are sufficient to not only drive allergic pathology, but also amplify adaptive Th2 cell responses6–9. These studies suggest that innate immune mechanisms may play a larger role in driving atopic inflammation than previously recognized.

Innate lymphoid cells (ILCs), while lacking antigen receptors generated by recombination activating gene proteins RAG1 and RAG2 (collectively called RAG), are the innate counterparts of T cells. For example, ILC2s mirror adaptive Th2 cells in their developmental requirements, cytokine profiles, and effector functions10. Unlike classical T cells, ILC2s are concentrated at barrier surfaces to rapidly respond to microbial and environmental stimuli. ILC2s are key mediators of inflammatory skin conditions like AD11–13. Indeed, in murine models of AD-like disease, type 2 skin inflammation can still occur despite the absence of adaptive T cells, but is further reduced after depletion of ILC2s12,13. Furthermore, recent studies have shown that ILC2s harbor non-redundant functions in the presence of the adaptive immune system in the setting of anti-helminth immunity14,15. These findings suggest that ILC2 dysfunction may also uniquely contribute to the pathogenesis of atopic diseases, independent of adaptive immunity. However, the cell-intrinsic mechanisms that drive ILC2 dysregulation remain poorly understood.

ILC2s were originally discovered due to their capacity to orchestrate multiple allergic pathologies in immunocompromised mice, most notably in RAG-deficient mice that lack T and B cells16–22. These discoveries fundamentally redefined our understanding of allergic diseases and placed a major focus on ILC2s as potential drivers of human allergic disease. However, despite ILC2s not requiring RAG expression for their development, fate mapping studies in mice have demonstrated that up to 60% of ILC2s have historically expressed RAG1 during development23,24. Although previous work has described roles of RAG beyond antigen receptor recombination in developing T and B cells25–27 and NK cells24, how this developmental expression of RAG impacts ILC2s remains unclear.

By directly comparing RAG-deficient and RAG-sufficient mice, we unexpectedly found enhanced AD-like disease in RAG-deficient mice, despite the lack of adaptive lymphocytes to contribute to AD-like inflammation. Using splenocyte replenishment and bone marrow chimeras, we show that RAG suppresses ILC2 activation and expansion in a cell-intrinsic manner. Employing a RAG1-lineage reporter mouse line, we performed simultaneous single-cell multiomic RNA and ATAC sequencing to show that RAG fate-mapped ILC2s display unique transcriptional and epigenomic alterations consistent with the suppression of effector cytokine production. Collectively, our studies reveal evolutionarily conserved regulatory functions of RAG within innate lymphocytes, extending beyond the generation of antigen receptors in adaptive lymphocytes.

Results

RAG deficiency leads to expansion and activation of ILC2s

AD-like disease can be elicited in the skin of mice with repeated application of the topical vitamin D analog calcipotriol (MC903)28. Although it has been previously demonstrated that MC903 can induce AD-like disease in RAG-deficient mice that lack T and B cells, in part via ILC2 activation12,13, the relative contributions of ILC2s and the adaptive lymphocyte compartment have not been rigorously evaluated. We hypothesized that the presence of Th2 cells, in addition to ILC2s, would lead to enhanced AD-like disease in an additive fashion. In testing this, we evaluated both RAG1-sufficient wild-type (WT) mice and RAG1-deficient Rag1-/- mice in the setting of AD-like disease (Fig. 1A). Unexpectedly, we observed that Rag1-/- mice developed increased ear skin thickness (Fig. 1B) and increased absolute numbers and proportion of ILC2s in the skin-draining lymph nodes (sdLNs) compared to control WT mice (Fig. 1C,D; S1A,B). Furthermore, a larger proportion of ILC2s from Rag1-/- mice exhibited production of both IL-5 (Fig. 1E, S1C) and IL-13 (Fig. 1F, S1C). Our findings indicated that RAG1 deficiency results in paradoxically worse AD-like disease in association with enhanced ILC2 expansion and activation.

RAG deficiency leads to expansion and activation of ILC2s during inflammation and at steady state.

A) Experimental schematic of AD-like disease. WT (Control) B6 mice or Rag1-/- mice treated topically to the inner surface of each ear with 2 nmol MC903 in 10 μL ethanol vehicle daily for 7 days develop AD-like inflammation. B) Ear thickness measured daily in AD-like inflammation. Data representative of at least 2 independent experiments, 5 mice/group. ** P < 0.01 by 2-way ANOVA with Sidak’s multiple comparisons test, day 7. C) Total number ILC2s normalized to 105 live cells and D) ILC2 proportion of CD90+, Lin- cells (Lin- defined as CD3-, CD5-, CD11b-, CD11c-, CD19-, NK1.1-, and FcεR1-) determined to be ILC2s (IL-33R+) in skin-draining lymph nodes (sdLN) from WT or Rag1-/- mice with AD-like ear inflammation. Percent ILC2 from sdLN in AD-like disease following PMA/iono stimulation positive for E) IL-5 or F) IL-13 staining. G) Schematic of steady state analysis of sdLN from WT (Control) or Rag1-/- mice. H) Total number ILC2s normalized to 105 live cells and I) ILC2 proportion of steady state sdLN CD90+, Lin- cells determined to be ILC2s as in (C,D). Percent ILC2 from sdLN in steady state following PMA/iono stimulation positive for J) IL-5 or K) IL-13 staining. C-F; H-K) Data representative of at least 2 independent experiments, 4-5 mice/group * P < 0.05, ** P < 0.01, *** P < 0.001 by two-tailed Welch’s t test. All data represented as mean with standard deviation.

To determine whether this phenomenon was specific to AD-like pathological conditions, we next examined the sdLNs in Rag1-/- and lymphocyte-sufficient Rag+/- littermate control mice in the absence of disease (Fig. 1G). We found that the absolute number and frequency of ILC2s was increased at steady state in Rag1-/- sdLNs (Fig. 1H,I) and that a higher proportion of these ILC2s produced both IL-5 (Fig. 1J) and IL-13 (Fig. 1K) compared to WT controls following ex vivo stimulation and intracellular cytokine staining. The RAG recombinase requires both RAG1 and RAG2 components to successfully rearrange a functional antigen receptor in adaptive lymphocytes29. Thus, to test whether our findings are specific to RAG1, or related to function of the overall RAG complex, we similarly examined the steady-state profile of ILC2s in Rag2-/- mice (Fig. S2A). Deficiency of RAG2 led to an expansion of ILC2s in the sdLNs (Fig. S2B,C) and increased proportions of ILC2s expressing IL-5 (Fig. S2D) and IL-13 (Fig. S2E) similar to Rag1-/-mice. Collectively, these findings suggest that the RAG recombinase modulates ILC2 function at steady state and during type 2 inflammation. However, whether the hyperactive ILC2 phenotype is due to a cell-intrinsic process or simply due to the absence of T and B cells was unclear.

ILC2 suppression by RAG is cell intrinsic

Given the importance of the adaptive lymphocyte compartment in shaping the secondary lymphoid organ repertoire, we next wanted to examine whether the presence of adaptive lymphocytes could restore ILC2 homeostasis in RAG-deficient mice. To test this, we created splenocyte chimera mice by reconstituting both Rag1-/-and control WT mice with splenocytes containing T and B cells from WT donor mice (Fig. 2A). We first assessed the overall level of immune reconstitution in the recipient mice and found fully restored proportions of CD4+ (Fig. S3A) and CD8+ (Fig. S3B) T cells in the spleens of recipient Rag1-/- mice, although B cell numbers remained significantly lower than in WT mice (Fig. S3C). Upon induction of AD-like disease, we found that the Rag1-/-mice still exhibited increased ear skin thickness (Fig. 2B), enhanced expansion of ILC2s (Fig. 2C,D), and increased proportions of ILC2s expressing IL-5 (Fig. 2E) and IL-13 (Fig. 2F) in the sdLNs. Interestingly, we found significantly higher proportions of eosinophils in the spleens of Rag1-/- recipient mice (Fig. S3D), possibly reflecting the increased IL-5 production we observed from ILC2s. These findings indicate that the mere introduction of exogenous T and B cells is not sufficient to suppress ILC2 dysregulation in the setting of RAG deficiency.

Homeostatic expansion and activation of RAG-deficient ILC2s is cell intrinsic

A) Experimental schematic of AD-like disease in splenocyte chimera experiment. WT B6 or Rag1-/-mice received WT splenocytes and developed AD-like inflammation after subsequent topical treatment with 2 nmol MC903 in 10 μL ethanol vehicle to each ear daily for 10 days. B) Ear thickness measured daily in AD-like inflammation. Data representative of 2 independent experiments, 4-5 mice per group. **** P <0.0001 by 2-way ANOVA with Sidak’s multiple comparisons test, day 10. C) Total number ILC2s normalized to 105 live cells and D) ILC2 proportion of CD90+, Lin- cells (Lin- defined as CD3-, CD5-, CD11b-, CD11c-, CD19-, NK1.1-, and FcεR1-) determined to be ILC2s (IL-33R+). Percent ILC2 from sdLN in splenocyte chimera mice with AD-like disease after PMA/iono stimulation positive for E) IL-5 or F) IL-13 staining. G) Schematic of bone marrow chimera experiment. Equal quantities of bone marrow cells from Rag1-/- (CD45.2, CD90.2 - orange) and WT (CD45.2, CD90.1 - blue) C57Bl/6J donor mice were used to reconstitute the immune systems of irradiated recipient WT (CD45.1 - black) C57Bl/6J mice. H) Proportion of donor (CD45.2+) ILC2 defined as in (C) in sdLN by donor source (CD90.1+ - WT, CD90.2- - Rag1-/-). Proportion of Lin- ILCs by donor source positive for I) IL-5 and J) IL-13 following PMA/iono stimulation and cytokine staining. C-F) Data representative of at least 2 independent experiments, 4-5 mice per group. ** P < 0.01, **** P < 0.0001 by two-tailed Welch’s t test. H-J) Data representative of at least 2 independent experiments, 4-5 mice per group. ** P < 0.01 by two-tailed ratio means paired t test. All data represented as mean with standard deviation.

To further test whether this phenotype is mediated by cell-intrinsic RAG expression, we next generated mixed bone marrow (BM) chimeras. We harvested BM from congenic CD90.1+ WT and CD90.2+ Rag1-/- donor mice on a CD45.2+ background and reconstituted sub-lethally irradiated CD45.1+ congenic WT recipients with a 50:50 mixture of WT:Rag1-/- BM (Fig. 2G). After confirming reconstitution of donor immune cells in the sdLN (Fig. S4A,D), we examined the frequency and activity of ILC2s in the sdLNs based on whether they originated from WT (CD90.1+) or Rag1-/-(CD90.2+) donors (Fig. S4B,C). Strikingly, of the total donor ILC2s, the majority were derived from Rag1-/- donors (Fig. 2F,G). This was not due to differences in overall donor reconstitution, since measuring all Lin- cells revealed WT donor cells outnumbered those from Rag1-/- donors (Fig. S4E). Of the total IL-5- (Fig. 2H,I,K) and IL-13-expressing (Fig. 2 H,J,L) ILCs, the majority were also derived from Rag1-/- donors. Taken together, these data suggest that cell-intrinsic RAG activity in ILC2s may limit their capacity to expand and become activated.

A history of RAG expression marks a subpopulation of ILC2s in the skin draining lymph node

In contrast to resting naïve T cells, ILC2s resemble activated Th2 cells at steady state based on their transcriptomic and epigenomic profiles30,31. While both T cells and ILC2s exhibit historical RAG expression23, they do not actively express the protein in their mature state32. Taken together, these findings provoke the hypothesis that ILC2s are regulated by RAG early in development to imprint alterations that influence their activity as mature cells. To distinguish ILC2s as either having a history of RAG expression or not, we utilized a RAG lineage tracing system, whereby a Rag1Cremouse was crossed to a reporter mouse expressing tandem dimer red fluorescent protein (tdRFP) in a Cre-dependent manner from the Rosa26 locus (Fig. 3A)24,33. This system allowed us to compare RAG-experienced (RAGexp) and RAG-naïve (RAGnaïve) lymphoid cells, including ILC2s, simultaneously originating from the same immunocompetent host, thus removing confounders inherent in knockout and chimera experiments. Analysis of sdLN from the reporter mice revealed that nearly all CD4+ T cells, CD8+ T cells, and B220+ B cells expressed tdRFP (positive history of Cre expression from the Rag1 locus), consistent with the known requirement of RAG expression for their development (Fig. 3B-D,G). We also examined NK cells, since certain subsets of NK cells are known to express RAG during their development24, and we observed that 60% of NK cells were tdRFP+, similar to previous findings (Fig. 3E,G)24. In the ILC2 population of the sdLN, around 50% were tdRFP+ (Fig. 3F,G), similar to proportions of RAG fate-mapped ILC2s previously observed in the fat24 and lung23,24. These findings demonstrate that there are heterogeneous populations of ILC2s marked by differential tdRFP+ fate mapping. Importantly, this provided us with the possibility to profile these different ILC2 subsets based on Rag1Cre-activated expression of tdRFP.

A history of RAG expression marks a population of ILC2s in the sdLN.

A) Schematic of RAG fate mapping in the lymphoid cell compartment using reporter mice expressing Cre-inducible tandem dimer red fluorescent protein (tdRFP) from the Rosa26 locus crossed to mice expressing Cre recombinase from the Rag1 locus. B-F) Histograms of tdRFP signal in CD45+ sdLN cells by cell type for B) CD4+ T cells (B220-, CD3+, CD4+), C) CD8+ T cells (B220-, CD3+, CD8+), D) B cells (MHCII+, B220+), E) NK cells (B220-, CD3-, CD4-, CD8-, CD49b+, NK1.1+), F) ILC2s (B220-, CD3-, CD4-, CD8-, CD49b-, NK1.1-, CD11b-, CD11c-, SiglecF-, CD90+, KLRG1+ or ICOS+ or IL-33R+), F) quantification of tdRFP+ proportion of each cell type.

Multiomic profiling enhances the detection of rare tissue-specific ILC2s

Transient RAG expression early in lymphoid development leads to well-characterized, durable effects on B and T cell development and function mainly through successful genomic rearrangement of antigen receptors. Yet, our data indicate that RAG expression also imprints phenotypic changes on ILC2s, which can develop independently of functional antigen receptors, provoking the hypothesis that RAG expression may affect broader epigenomic and transcriptional programs. Furthermore, our data indicate that the impact of RAG on ILC2 function has implications for AD-like skin inflammation, suggesting a persistent effect that modulates phenotypes of type 2 inflammation.

To test these hypotheses, we performed combined single-nuclei RNA sequencing (snRNA-seq) and ATAC sequencing (snATAC-seq) of sdLN cells from RAG fate-mapped mice at steady state and in the setting of AD-like disease (Fig. S5A,B). Because fluorescence-activated cell sorting (FACS) can cause physical stress, cell loss, and contamination, which can introduce unwanted perturbations in target cells, instead we utilized gentle initial negative selection by magnetic activated cell sorting (MACS) to remove most B and T cells and monocytes prior to sequencing (Fig. S5A). This allowed us to enrich for innate immune cell populations prior to sequencing. Further, we used the gene encoding tdRFP as a barcode to differentiate between RAGexp (RAG fate map-positive) and RAGnaïve (RAG fate map-negative) ILC2s at the single-cell level (Fig. 5A). The multiomic data was analyzed using recently developed pipelines in Cell Ranger, Seurat34–36, and Signac37, and sequenced cells were further filtered computationally to enrich for ILCs, as in previous studies (see methods)38.

In addition to gene expression (GEX) information derived from snRNA-seq (Fig. 4A), we calculated a “gene activity” (GA) score based on chromatin accessibility at gene loci (Fig. 4B) from the corresponding snATAC-seq dataset37. Clustering the cells with each data subset alone and in combination using weighted nearest neighbor (WNN) analysis, we identified six clusters (Fig. 4C) that demonstrated consistent differences in cellular markers based on both metrics of GEX and GA (Fig. 4D,E; Table S1). Additionally, top markers for each cluster clearly differentiated each cell type (Fig. 4F). Despite successful ILC2 enrichment via MACS depletion for lineage markers and computational filtering (see methods), our data set included non-ILC2 populations determined by gene expression to be T cells, dendritic cells, B cells, and NK cells (Fig. 4A,F; Table S1), which allowed for broader multidimensional comparisons while studying this ILC2-enriched data set.

Multiomic analysis of ILC2s through single nuclei sequencing of the sdLN.

A) Schematic of the gene expression (GEX) assay derived from snRNA-seq data. B) Schematic of the gene activity (GA) assay, representing estimated transcription scores derived from snATAC-seq data using Signac. C) UMAP visualizations of independent analyses of RNA-seq and ATAC-seq data for 2034 sdLN cells after dimensional reduction and clustering combined using weighted nearest neighbor (WNN) analysis in Seurat. Cluster identities are color coded consistently throughout the following panels. Heatmaps of D) top 25 GEX marker genes and E) top 25 GA marker genes identified for each cluster. See Table S1 for full lists of genes. F) Dotplots comparing selected marker genes for each cluster between the GEX and GA assays, with emphasis on known cell type-specific markers. G) Schematic of differentially accessible (DA) chromatin assay, which finds the nearest gene to any peak calculated to be differentially open in a particular cell cluster. See Table S2 for full lists of top 25 DA cluster markers. H) Overlap of top 100 markers for the ILC2 cluster from the GEX, GA, and DA assays. See Table S3 for top 100 DA peaks and distances to nearest genes and Table S4 for full list of top 100 ILC2 markers. I) Selected genes from the ILC2 gene set for each assay individually and for overlaps.

To further complement the GEX and GA assays, we utilized another method of detecting cell-specific marker genes, whereby chromatin regions that are differentially accessible (DA), or open, in each cluster could be linked by their physical proximity to specific genes (Fig. 4G, S6; Table S2; see methods). Comparing the top 100 GEX, GA, and DA markers in the ILC2 population, we identified a multiomic ILC2 signature of 235 unique genes (Fig. 4H; Table S3, 4). While there was some overlap between each respective set, the multiomic approach enabled more extensive identification of an ILC2 gene program than through either snRNA-seq or snATAC-seq alone. The analysis revealed a variety of canonical ILC2-associated genes specific to the ILC2 cluster (Fig. 4I, S6,7) including the ILC2-activating cell surface receptors Icos39,40, Il2ra41,42, Il18r138,43,44, Nmur14,15,45–47, and Il1rl1 (encoding the receptor for IL-33)11,12,48,49, as well as transcription factors such as Gata319,50–53, Bcl11b54–57, Maf19,58,59, Ets160, and Rora38,51,61–63, all previously shown to be important for ILC2 development and/or function.

Expression of some secreted proteins can be difficult to capture in droplet-based snRNA-seq experiments due to low transcript levels and relatively shallow sequencing depth. With the addition of the complementary GA and DA assays from snATAC-seq, our analysis identified Il5 (Fig. 4I, S6B), a canonical ILC2 cytokine, in the DA assay, while in the GA assay we found Bmp764,65, which has been shown to be secreted by ILC2s to influence browning of adipose tissue. Additionally, we identified the secreted chemokine Ccl1 as an ILC2 marker gene66–69, which along with its cognate receptor Ccr8 (also an ILC2 marker in our analysis)44,67,70, participates in a feed-forward circuit to drive ILC2 recruitment and expansion71. Thus, our findings demonstrate how genetic barcoding, combining transcriptomic and epigenomic analyses, and cross-validation across many published studies can yield new insights while providing internal control measures to elevate the rigor, robustness, and confidence of identifying gene signatures of rare populations such as ILC2s at the single-cell level.

ILC2s with a history of RAG expression are epigenomically suppressed

As noted above, barcoding the ILC2s afforded the opportunity to transcriptionally and epigenomically profile ILC2s under identical developmental conditions by dividing the ILC2 cluster into RAGexp and RAGnaïve populations. We hypothesized that RAGexp ILC2s would have a distinct transcriptional profile compared to ILC2s without any history of RAG expression. To test this, we calculated differentially expressed genes (DEGs) for the ILC2 cluster by Rag1 fate-map status. Genes with higher expression in RAGexp cells relative to RAGnaïve cells had positive fold change values, and vice versa, with genes relatively increased in RAGnaïve cells having negative values (Fig. S8A, Table S5). Using gene set enrichment analysis (GSEA)72,73 on the ranked list of DEGs, we found that gene sets generally representing lymphocyte activation and differentiation were suppressed in RAGexp ILC2s compared to RAGnaïve ILC2s (Fig. S8B,C, Table S6), consistent with our previous observations that ILC2s are expanded and more activated in RAG-deficient mice relative to WT mice.

We next employed newly described methodologies37,74 that quantify associations between open chromatin peaks and the expression of nearby genes to describe the functional regulomes of both RAGexp and RAGnaïve ILC2s (Fig. 5B). In this analysis, each ATAC peak can be linked to multiple genes, and each gene to multiple peaks, generating a list of “gene-to-peak links” or GPLs (see methods). For each gene, we interpreted the number of corresponding GPLs as a quantitative representation of the regulome activity for that gene. Considering RAGexp and RAGnaïve cells as two separate populations, we generated two lists of GPLs (Table S7) defining functional regulomes for each population. We focused our analysis on the functional regulomes of ILC2s by filtering the GPL lists for the 235 unique ILC2 genes identified in our multiomic gene set (Fig. 4H; Table S8), then calculated the difference in GPLs between RAGexp and RAGnaïve cells for each gene and sorted them; genes displaying greater numbers of GPLs in the RAGnaïve population are at the top, and genes with more GPLs in the RAGexp population are at the bottom (Fig. 5C; Table S9).

A history of RAG expression imprints transcriptomic and epigenomic modulation of ILC2 gene programs.

A) Schematic of transcriptional RAG fate mapping. Sequenced cells from the RAG fate map mouse (see Figure 3A) transcribe tdRFP only after Cre is expressed from the Rag1 locus. Cells were assigned as either having a history of RAG expression (RAGexp - tomato red) or not (RAGnaïve - dark gray) based on detection of tdRFP transcript in the RNA-seq data. B) Schematic of mapping gene to peak links (GPLs). The LinkPeaks function of Signac (see methods) calculates significant correlations between open chromatin at defined peaks (teal bars) and nearby gene expression. These links represent inferred epigenomic-transcriptomic regulation, or "regulomes" based on the correlated snRNA- and snATAC-sequencing data. After calculating GPLs separately for each population (gray for RAGnaïve and tomato red for RAGexp), GPLs found in only one group, but not the other, can be identified (teal boxes). The difference in GPLs based on RAG experience for any given gene (e.g. Gene X) can be visualized on a bar graph, with the number of GPLs for RAGnaïve (gray - left) and RAGexp (red - right) plotted, with the difference overlaid as a black bar. C) GPLs calculated as in (B) for the multiomic ILC2 gene set identified in Figure 4H and Table S4). All identified GPLs are listed in Table S7, while ILC2 GPLs are listed in Table S8. Genes are sorted from more links identified in the RAGnaïve population at top to more links identified in the RAGexp population at bottom. Select genes are labeled. Full ranked list by difference in GPLs is available in Table S9.

We found that the ILC2 marker genes segregating toward the top of this list, corresponding to enhanced epigenomic activity in the RAGnaïve cells, tended to be genes previously identified to play positive roles in development, expansion, and activation of lymphoid cells. These included transcriptional regulators such as Tox75–78, Rora38,51,61–63, Maf19,58,59, and Gata319,50–52, which are involved in early differentiation of both ILCs and lymphocytes (Fig. S9A). Indeed, epigenetic activation of the Gata3 locus is recognized to play a critical role in development of both ILC2s79 and Th2 cells80,81. Additionally, surface receptors known to drive ILC2 activation upon stimulation including Il18r138,43,44, Il1rl121,38,43,82,83, and Icos39,40 had increased functional regulome activity in RAGnaïve ILC2s. In contrast, genes with more GPLs in the RAGexp ILC2s tended to be associated with suppressive functions. For example, Ndfip1 (Fig. S9B) encodes a regulatory protein that enhances activity of the ubiquitin ligase ITCH to negatively regulate inflammation84,85 and has been associated with asthma risk in GWAS studies86. Dusp1 partially mediates glucocorticoid effects through its ability to negatively regulate inflammation87,88, is associated with eczema by GWAS89, and has recently been shown to mark an anti-inflammatory set of ILCs69. Last, Asxl1 encodes a tumor suppressor that inhibits clonal hematopoiesis through its epigenomic regulatory effects in both mice and humans90–93. Collectively, our GPL analysis stratifies the ILC2 gene signature based on RAG experience, where genes associated with ILC2 expansion and activation are poised in RAGnaïve cells, while genes associated with suppressive effects are poised in RAGexp cells.

We expanded our multiomic analysis to infer information about transcription factor (TF) activity from open chromatin regions in our snATAC data. We used the chromVAR package94, which finds known TF binding motifs in open chromatin regions in each cell, to identify TF motifs enriched in each cell cluster (Fig. S10A,B, Table S10). The enriched motifs were consistent with known functional roles of associated TFs in each cell type. For example, in the NK cell cluster we found enriched motifs recognized by the TFs EOMES and T-bet (encoded by Eomes and Tbx21, respectively), which are critical for development of NK cells95. A limitation of this analysis is that while TF motif accessibility can be inferred from open chromatin in snATAC data, which TFs are bound to the identified accessible sites is not known. We reasoned that complementary gene expression information from our multiomic data could mitigate this limitation in part by comparing accessibility of TF binding motifs to expression levels of corresponding TFs (Fig. S10C). Indeed, motifs for both RORα and RORγ (encoded by Rora and Rorg, respectively), which share a common DNA binding 5’-AGGTCA-3’ half site, have similar calculated accessibilities in both the ILC2 and NK cell clusters. Yet only Rora is expressed at appreciable levels, and only in ILC2s, consistent with its critical role in ILC2 development61,62. In contrast, ILC2 development is not dependent on Rorg expression, and neither RORα nor RORγ plays a major role in NK cells. Taken together, this analysis confirms the known role of Rora in ILC2s and highlights how matched multiomic chromatin accessibility and gene expression data can clarify ambiguities inherent in TF enrichment analyses.

The broad effects of RAG expression on ILC2 transcriptional regulomes we observed (Fig. 5C) led us to hypothesize that distinct cohorts of TFs may contribute to the differences observed between RAGnaïve and RAGexp ILC2s. To test this hypothesis, we analyzed the open chromatin regions in GPLs unique to each RAG fate mapped ILC2 population using the FindMotifs function in Signac37, which returns a ranked list of enriched motifs corrected for background presence of each motif in all cells. In both RAGnaïve and RAGexp ILC2s, we identified enriched TF motifs (Fig. S10D, Table S11) that are GC rich regions recognized by a large family of C2H2 zinc finger TFs, particularly the Krüppel-like factors (KLFs), which are well-established as key regulators of lymphocyte development96,97. Given the strong sequence similarities of the identified TF motifs, we turned to the matched gene expression data to clarify which TFs may be available to engage the accessible binding sites. Of the eleven unique TFs identified, only six were detected in the gene expression assay (Fig. S10E). We observed much higher expression of Klf2, Klf6, and Klf12 in Ragexp ILC2s compared to RAGnaïve ILC2s (Fig. S10E). Notably, all three of these TFs have been associated with reduced cellular proliferation and/or activation98–100. Klf2 expression plays a key role in T cell quiescence101,102, and both Klf2 and Klf6 were recently identified as markers of “quiescent-like” skin resident ILCs69. In contrast, although detected in a smaller fraction of cells in our data, we found Klf7 expression was higher in RAGnaïve ILC2s compared to Ragexp ILC2s (Fig. S10E). Increased expression of Klf7 has been shown to enhance survival of early thymocytes and is a predictor of poor outcome in acute lymphoblastic leukemia103. Collectively, these findings link the relatively activated or suppressed epigenomic and transcriptomic states of RAGnaïve and Ragexp ILC2s, respectively, to distinct cohorts of homeostatic TFs.

A history of RAG expression modulates ILC2 epigenomes at steady state and in AD-like inflammation

Although our GPL and TF analyses revealed a suppressive effect of RAG expression on ILC2 gene programs, we did not account for the additional variable of disease state in the initial analysis. To test whether RAG expression promotes a suppressive epigenomic program in ILC2s that is durable in the setting of inflammation, we first recalculated GPLs after splitting our dataset by both history of RAG expression (naïve vs. experienced) and disease (steady state vs. AD-like disease) to yield four lists of GPLs (Fig 6A; Table S12). When we examined the intersection, or overlap, of peaks from ILC2 GPLs (Table S13), several notable patterns emerged (Fig. 6B). First, the largest set of peaks was shared by all RAGnaïve cells (gray bar), regardless of disease state, with the next largest peak sets belonging to either steady state or AD-like disease in the RAGnaïve cells. Second, there was a large set of peaks shared by all RAGexp cells (red bar). Third, the intersections corresponding to disease states (steady state – yellow, AD-like disease – dark red), had relatively few unique peaks. These findings suggest that early exposure to RAG expression plays a larger role in modulating the epigenomic signature of the ILC2 gene program than exposure to disease. To confirm that the patterns we observed represent a specific effect of RAG expression on the ILC2 gene program, we performed the same analysis on GPL peaks for all genes in the dataset (Fig. S11). In contrast to the ILC2 gene set, the majority of GPL peaks for all genes was shared among all cell populations, consistent with epigenomic regulation of most genes being minimally affected by either RAG expression or AD-like disease. Last, in the ILC2 gene set analysis, we noted a set of poised peaks shared by all RAGnaive cells and RAGexp cells in the setting of AD-like disease, but not with RAGexp cells at steady state (blue bar, Fig. 6B). We reasoned that this condition might capture some genomic loci that are suppressed by a history of RAG expression at steady state but are induced during inflammation.

A history of RAG expression broadly influences ILC2 genes at steady state and in AD-like inflammation.

A) Schematic of the process to determine contribution of RAG fate map and disease states to GPLs for subsequent intersection analyses. GPLs were first calculated for all indictated cells, regardless of disease state or fate map. Cells were then split, first by RAG fate map (RAGexp and RAGnaïve), and again by disease state (SS - steady state, AD - AD-like inflammation). GPLs were recalculated for each split sample and matched back to the original set of total GPLs. B) UpSet plot visualizing intersections of peaks identified from ILC2 GPLs for split samples. Each row represents one of the four sets, and each column corresponds to an intersection of one or more sets (see methods). See Table S12 for full list of GPLs for all genes. Table S13 lists total and ILC2 peaks used for intersection analyses in each of the four sets. Columns identifying key intersections are color coded by the corresponding RAG fate map or treatment groups. The blue column indicates the intersection of peaks from RAGnaïve cells and peaks induced by AD-like disease in RAGexp cells. C) Top genes with the most AD-like disease-induced peaks. Peaks from the intersection between RAGnaïve cells and inflamed RAGexp cells were identified in corresponding GPLs, and genes were ranked by number of linked peaks identified. See Table S14 for full list of ranked genes and associated GPLs. Open chromatin in the ILC2 cell cluster split by disease state and then by RAG fate map for the genomic loci of C) Ccr6 and D) Rora.

Thus, we next quantified and sorted these GPLs to generate a list of genes with the most peaks "induced" during AD-like disease (Fig. 6C, Table S14). Among the identified genes, we selected Rora (Fig. 6C) and Ccr6 (Fig. 6D) to examine more closely for evidence of epigenomic activation in AD-like disease, given the role of these genes in ILC2 expansion61,62 and homing to sites of inflammation44,104, respectively. For both genes, we observed more widespread open chromatin over the genomic region in the RAGnaïve cells compared to the RAGexp cells, but this difference was partially abolished by increased open chromatin in AD-like disease in the RAGexp cells. Taken together, our analysis reveals that a history of RAG expression selectively modulates activity of ILC2 gene programs across both steady state and during AD-like inflammation, while some programs are more evident at steady state given the uniquely poised nature of ILC2s.

RAG suppresses the Th2 locus

Our functional data demonstrate a role for RAG expression in regulating ILC2 development and activation, including limiting proportions of IL5+ and IL-13+ ILC2s at steady state and in AD-like disease. Prior work identified epigenomic priming in ILC2s early in development at the Th2 locus (comprised of the Il4, Il13, Rad50, and Il5 gene loci) to enable rapid transcriptional responses during inflammation30. Thus, we hypothesized that RAG promotes the functional observations in ILC2s by suppressing the establishment of an active regulome at the Th2 locus. To test this hypothesis, we analyzed the Th2 locus in our multiomic data in greater detail. Using a similar strategy to our analysis of ILC2 marker genes, we calculated the number of GPLs in the RAGexp and RAGnaïve cells, respectively, for the genes in the Th2 locus. (Fig. 7A, Table S7). We found many GPLs associated with the four Th2 locus genes, including significant crosstalk between these genes, similar to previous observations (Fig. 7A)105–107. Importantly, we identified fewer GPLs in the RAGexp cells, particularly for the Il5 and Il13 loci (Fig. 7A). As in our analysis of the ILC2 marker GPLs, we quantified the differences based on RAG fate mapping and found that all genes in this locus had increased GPLs in RAGnaïve cells relative to RAGexp cells (Fig. 6B; Table S15). We also applied the same analysis strategy that identified TFs potentially mediating observed differences between RAGexp and RAGnaïve ILC2s (Fig. S10D,E) specifically to the four Th2 locus genes. Given the limited size of genomic regions (and thus open chromatin peaks) analyzed in the Th2 locus compared to all ILC2 genes, we found overall fewer enriched motifs. Strikingly, significant enrichment of TF motifs was only present in unique peaks from RAGnaïve ILC2s, while no TF motifs met the cutoff in RAGexp cells (Fig S12A, Table S16). These motifs primarily contained the canonical 5’-(A/T)GATA(A/G)-3’ binding site recognized by the GATA family of zinc finger TFs108. When we compared enriched motifs in open chromatin to gene expression of the corresponding TFs, only Gata3 was expressed at appreciable levels (Fig S12B). Critically, Gata3 expression was higher in RAGnaïve compared to RAGexp cells, consistent with our previous analyses of the ILC2 gene regulomes (Fig. 5C). Collectively, our data confirm the established role of GATA3 in mediating activation of the Th2 locus50 and are consistent with a role for RAG expression in suppressing the type 2 regulome at the Th2 locus.

RAG suppresses the Th2 locus.

A) Coverage plot of the Th2 genomic locus. Open chromatin in the ILC2 cluster for each Rag1 fate-mapped state is shown on top, and corresponding peaks (teal) and gene to peaks links (GPLs) are shown below for the RAGnaïve sample (gray) and the RAGexp sample (tomato red). Only GPLs that fit in the coverage window are shown. B) All GPLs identified in each fate map state for the Th2 locus genes Il4, Il13, Rad50, and Il5. See Table S15 for full list of Th2 GPLs. The number of GPLs for each gene is shown on the left in gray for RAGnaïve and on the right in tomato red for RAGexp. The difference is superimposed in black, and genes are sorted from more GPLs identified in RAGnaïve at top to more links identified in RAGexp at bottom. C) UpSet plot of intersections of peaks identified from Th2 locus GPLs. GPLs were recalculated, this time for samples separately by both RAG fate map status (RAGexp and RAGnaïve) and disease (SS - steady state, AD - AD-like inflammation). Each row represents one of the four sets of peaks, and each column corresponds to an intersection of one or more sets (see methods). See Table S13 for full list of peaks from GPLs for all genes, including Th2 genes, in each of the four sets. Columns identifying key intersections are color coded by the corresponding RAG fate map or disease groups. The blue column indicates the intersection of peaks from RAGnaïve cells and peaks induced in AD-like disease in RAGexp cells. (D) Th2 genes sorted by number of AD-like disease-induced peaks. Peaks induced by AD-like disease were identified in corresponding GPLs, and genes were ranked by frequency of links to induced peaks (representation in identified GPLs). See Table S17 for full list of ranked Th2 locus genes and associated GPLs. E) Open chromatin tracks, split by disease (beige box – steady state; maroon box – AD-like disease) and by RAG fate map ( RAGnaïve - gray, RAGexp - red) for Il13.

We next considered the additional effects of AD-like inflammation on the Th2 epigenomic regulome using the same approach we used to analyze the ILC2 gene set in Fig. 6. Again, we found the largest set of peaks was shared by the RAGnaïve cells, regardless of disease state, with the next largest peak sets belonging to either steady state or AD-like disease in the RAGnaïve cells (Fig. 7C). Furthermore, there was a large proportion of peaks shared by both RAGexp cells, consistent with a major contribution of a history of RAG expression to epigenomic modulation of the Th2 regulome. To quantify the potential effect of AD-like inflammation on reversing RAG-mediated suppression of Th2 locus genes, we mapped the 14 peaks shared by RAGnaive cells and RAGexp cells in the setting of AD-like inflammation (i.e. only suppressed in RAGnaive cells at steady state) back to the Th2 genes via their respective GPLs (Fig. 7C - blue bar, Table S17). Interestingly, Il13, which was not identified as a top ILC2 marker in our earlier analyses, had the highest number of linked peaks associated with potential induction in AD-like disease (Fig. 7D). When we examined the Il13 locus in the ILC2 cluster more closely, we found more widespread open chromatin in the RAGnaïve cells compared to the RAGexp cells (Fig. 7E). However, in the AD-like disease sample, the RAGexp cells displayed increased open chromatin relative to the steady state, consistent with induction in the setting of inflammation, like our earlier findings for ILC2 genes such as Ccr6 (Fig. 6E). Taken together, our functional data and multiomic analyses demonstrate a role for RAG expression in modulating genes critical for ILC2 development and function, including the key type 2 cytokines expressed from the Th2 locus.

Discussion

RAG recombinases evolved nearly 500 million years ago from endogenous transposons, crucially enabling antigen receptor rearrangement and emergence of the adaptive immune cell lineages present in all modern vertebrates29,109. Indeed, RAG deficiency leads to a complete lack of B and T lymphocytes, manifesting clinically as severe combined immunodeficiency (SCID)110–112. However, fate mapping studies have shown that multiple mature immune cell populations other than adaptive B and T lymphocytes have a history of RAG expression23,24,33,113,114. More recent studies by Karo et al found that RAG expression during NK cell development influences multiple cellular functions including antitumor cytotoxicity, cell proliferation, and survival24. Yet whether RAG modulates cellular functions of innate immune cell populations other than NK cells remains poorly understood. Here, using RAG-deficient mice, RAG fate mapping mice, and multiomic analyses, we report that RAG suppresses developmental and effector functions of ILC2s.

Our functional data in RAG-deficient mice demonstrate that populations of ILC2s producing the type 2 cytokines IL-5 and IL-13 preferentially expand in the absence of a history of RAG expression. This implies a specific role for RAG in developmental repression of ILC2s. Building on this, our multiomic RAG fate mapping analyses of ILC2 gene programs demonstrate extensive epigenomic differences between RAGexp and RAGnaïve cells. We found RAG-associated epigenomic suppression at multiple functional levels, including cell surface receptors, key transcription factors, and the Th2 locus encoding the type 2 cytokine genes Il5 and Il13. Although RAG is only transiently expressed early in lymphoid development115, our data demonstrate that RAG expression can imprint durable effects on ILC2 gene programs to restrain their function.

Our observations imply that RAG expression may mark a developmentally distinct population of ILC2s. In adaptive lymphocytes, RAG expression in T cells is restricted to their time in the thymus. However, ILC2 populations have been observed in the thymus, provoking the hypothesis that thymic ILC2s may be uniquely high in expression of RAG63,116. Prior studies by Schneider et al have identified ILC2 populations in adult tissues that variably derive from expansion of fetal, postnatal, and adult populations68. Yet how RAG expression in ILC2s may be restricted spatially or temporally remains unknown. The mouse strains used in the fate mapping studies by Schneider et al would be incompatible with our RAG fate mapping mice. Thus, novel mouse strains enabling intersectional genetics to trace ILC2 ontogeny (e.g. CreER/lox for temporally restricted fate mapping and FlpO/frt for RAG fate mapping117) are needed to more precisely determine when and where RAG expression occurs during ILC2 development. Beyond steady state ontogeny, our data suggest a history of RAG expression also imprints suppressed proliferative and type 2 inflammatory functions on ILC2s in the setting of AD-like disease.

It is increasingly recognized that expression of effector molecules for both ILCs and their counterpart adaptive lymphocytes (e.g. IL-13 from ILC2s and Th2 cells) is governed by finely tuned transcriptomic and epigenomic regulomes30,118–120. ILCs tend to adopt these regulomes earlier in their development than T cells, and these “poised” regulatory elements are thought to underlie the ability of tissue-resident ILCs to rapidly respond to stimuli. In contrast, the regulomes of naïve T cells remain relatively inactive until stimulation. Given that T cells are uniformly RAG-experienced, our data provoke the hypothesis that RAGexp ILC2s adopt a phenotype closer to that of naive T cells and may require stronger stimuli than RAGnaïve ILC2s to become activated. Indeed, our analyses found the RAG-associated suppressive programs could be overcome in the setting of AD-like inflammation. Thus, sufficient RAG expression may mediate key events underlying the establishment and maintenance of functional regulomes not only in ILCs, but also T cells. How RAG might affect these changes, and whether they are independent of its enzymatic activity and/or antigen receptor recombination, remains to be elucidated.

Clinically, a link between enhanced type 2 immune activity and RAG dysfunction is well-established. Omenn Syndrome (OS) is a form of SCID characterized by exaggerated type 2 immune activation and typically arises in the setting of hypomorphic RAG gene mutations. Impaired antigen receptor rearrangement, with rare “leaky” recombination events, leads to expansion of autoreactive oligoclonal T cells, eosinophilia, and markedly elevated IgE121–125. Similar phenotypes have been observed in mice harboring RAG mutations analogous to those found in human patients with OS126,127. Notwithstanding these findings, the mechanisms underlying the propensity of oligoclonal T cells with hypomorphic RAG activity to preferentially develop into the Th2 subtype are unclear. Prior studies have found a role for regulatory T cells in controlling type 2 skewing of transferred T cells in RAG-deficient hosts, potentially explaining similar observations in patients with OS128. Our data provide an additional mechanism by which RAG dysfunction may lead to OS through loss of cell-intrinsic RAG-mediated suppression of type 2 cellular programs. Additionally, increased type 2 cytokine production from RAG-deficient ILC2s may, in trans, enhance expansion of the oligoclonal Th2 cell populations, IgE induction, and eosinophilia observed in RAG-deficient states like OS. However, whether other immune cell types with RAG dysfunction, such as ILCs, contribute to the pathogenesis of OS in humans has not been investigated.

Lymphoid acquisition of RAG activity may represent a newer evolutionary mechanism that fine tunes ancient innate immune cell programs in addition to enabling development of relatively newer antigen-specific adaptive immune cell populations. Independent of antigen receptor diversity, loss of this function may offer an explanation as to why oligoclonal T cells tend to expand and skew towards a Th2 cell phenotype in the setting of hypomorphic RAG function as in OS128. Further studies are needed to define whether the suppressive effects of RAG expression operate similarly in T and B cells. Although we demonstrate that this phenomenon is observed in ILC2s, whether hypomorphic RAG expression in bona fide Th2 cells not only results in oligoclonality but also loss of suppression of the Th2 locus independently of antigen receptor rearrangement remains an outstanding question. Indeed, during development of gene therapy strategies for RAG-deficient SCID, lower doses of wild type RAG transgene expression have been associated with development of OS-like conditions in transplanted RAG-deficient mice129–132.

A major limitation of our study is a focus on cutaneous type 2 inflammation, which stemmed from our initial observations in the MC903 mouse model of AD-like disease. Further, given the scarcity of skin-resident ILC2 populations, key functional investigations in our study such as cytokine production and multiomic sequencing were limited to the sdLN, as in prior studies13,133. However, ILC2s are recognized to have highly tissue-specific functions that extend much beyond inflammation to other processes including regeneration and metabolism. In addition to Il-5 and IL-13, ILC2s can produce other effector molecules such as acetylcholine, IL-9, methionine-enkephalin peptides, and amphiregulin, which modulate tissue responses across numerous organs134–141. Considering that the complexity of ILC2 biology may result in markedly divergent responses to RAG expression in other tissues and disease models, we thus restricted our initial studies to the skin, where we had strong molecular, cellular, and phenotypic outcomes. An implication of our findings in the skin is that RAG expression may modulate a variety of ILC2 functions in other tissues. Broader surveys of how RAG impacts ILC2 development and function in different tissues and disease states remains an exciting area of inquiry.

While we focused our multiomic analyses on ILC2s, it is likely that RAG may impact other ILC populations. For example, hyperactivation of intestinal ILC3s has been observed in Rag1-/- mice secondary to persistent phosphorylation of Signal Transducer and Activator of Transcription 3 (STAT3). Adoptive transfer of T regulatory cells rescued this phenotype, providing a cell-extrinsic mechanism for the observation of hyperactivated ILC3s in the setting of RAG deficiency142. However, our data supporting a cell-intrinsic role for RAG in ILC2s may offer additional mechanistic insight into the prior observations in ILC3s. We found that the regulome of Jak2, which encodes JAK2, an upstream activator of STAT3, was more activated in RAGnaïve ILC2s (Fig. 5C). Additionally, the regulome for Dusp1, which encodes dual specificity phosphatase 1 (DUSP1), was more activated in RAGexp ILC2s (Fig. 5C). While not implicated in directly dephosphorylating STAT3, a recent study found DUSP1 overexpression negatively regulated the JAK2/STAT3 pathway143. Notably, recent transcriptional profiling of skin ILCs identified a potential mechanism for skin ILC populations to transition to an ILC3-like phenotype69, but how this process is regulated remains poorly understood. Taken together, our data provoke compelling new hypotheses about cell-intrinsic functions of RAG that may be complementary, rather than contradictory, to prior observations in gut and skin ILC populations. Additionally, our studies provide a rationale to design novel reagents to enable more comprehensive studies on the role of RAG in multiple innate immune cell populations across different tissues and disease models.

Our observations are also limited by lack of a defined mechanism for how RAG expression imprints durable epigenomic and transcriptomic changes in ILC2s. The mechanisms by which RAG mediates VDJ recombination are well-defined, from the biochemical details of DNA-binding to the epigenomic accessibility of antigen receptor loci and timing of RAG expression29,144–146. Notwithstanding genomic stress24 or potential RAG dose effects129–132, how RAG expression might modulate broad developmental and functional lymphoid programs other than V(D)J recombination remains unclear. The RAG complex can bind both DNA and modified histones and has been observed to occupy thousands of sites across the genome27. Thus, RAG may directly influence open chromatin states or obscure transcription factor binding sites to alter ILC2 development and function. Notably, RAG preferentially binds near transcription start sites of open chromatin in mouse thymocytes and pre-B cells, although corresponding effects on gene expression were not observed27. Although canonical recombination sites are concentrated in the antigen receptor loci, cryptic recombination sites in other regions may be deleted or rearranged by RAG activity, altering transcriptional regulation of associated genes27. In contrast to developing B and T lymphocytes, the precise timing and location of RAG expression in ILC2s is not known. Thus, combined with the relative scarcity of ILC2s, conventional methods of chromatin immunoprecipitation to identify potential epigenomic regulatory mechanisms mediated by RAG expression may not be feasible in ILC2s or other rare cell populations. Instead, newer technologies such as self-reporting transposons147 could be adapted to trace the genomic footprint of RAG in cells at various stages of development and in various tissues independent of the constraint of concurrent RAG expression. Finally, through its E3 ubiquitin ligase activity29, RAG may influence immune signaling pathways independently of transcription altogether. Given that direct targeting of RAG would lead to unacceptable side effects, elucidating the mechanisms by which RAG imprints phenotypic changes beyond antigen receptor rearrangement is a critical next step in translating these findings to potential new therapies.

Our studies expand prior work implicating RAG in critical immune functions beyond antigen receptor rearrangement that is exclusive to adaptive lymphocytes. Further, we provide additional insights into why patients with OS exhibit atopic syndromes in the setting of adaptive lymphocyte deficiency. Future studies into mechanisms underlying these findings may lead to new therapeutic avenues for disorders such as atopic dermatitis, food allergy, and asthma.